---
output:
  html_document: default
  pdf_document:
    keep_tex: yes
---

#One-Way ANOVA {#chapter3}

```{r echo=F,warning=F,message=F}
set.seed(3234)
library(pander)
require(mosaic)
library(knitr)
knitr::opts_chunk$set(cache = F,fig.width = 5, fig.height = 3.5,fig.pos='ht',
                      message=F,warning=F)
options(show.signif.stars = FALSE)
```


```{r echo=F}
#Color Format
colFmt = function(x, color){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}
colFmtBold = function(x, color){
  outputFormat = opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{\\textbf{",x,"}}",sep="")
  else if(outputFormat == 'html')
    paste("<b><font color='",color,"'>",x,"</font></b>",sep="")
  else
    x
}
```

## Situation {#section3-1}

In Chapter \@ref(chapter2), tools for comparing the means of two groups
were considered. More generally, these methods are used for a quantitative
response and a categorical explanatory variable (group) which had two and only
two levels. The full prisoner rating data set actually contained three groups 
(Figure \@ref(fig:Figure3-1)) with *Beautiful*, *Average*, and *Unattractive* 
rated pictures randomly assigned to the subjects for sentence ratings. In a 
situation with more than two groups, we have two choices. First, we could rely 
on our two group comparisons, performing tests for every possible pair 
(*Beautiful* vs *Average*, *Beautiful* vs *Unattractive*, and *Average* vs
*Unattractive*). We spent Chapter \@ref(chapter2) doing inferences for differences
between *Average* and *Unattractive*. The other two comparisons would lead us to
initially end up with three p-values and no direct answer about our initial 
question of interest -- is there some overall difference in the average sentences
provided across the groups? In this chapter, we will learn a new method, called 
***Analysis of Variance***, ***ANOVA***, or sometimes ***AOV*** that directly assesses whether there is
evidence of some overall difference in the means among the groups. This version
of an ANOVA is called a ***One-Way ANOVA*** since there is just 
one^[In Chapter \@ref(chapter4), methods are discussed for when there are two
categorical explanatory variables that is called the Two-Way ANOVA and related 
ANOVA tests are used in Chapter \@ref(chapter8) for working with extensions of 
these models.] grouping variable. After we perform our One-Way ANOVA test for 
overall evidence of a difference, we will revisit the comparisons similar to those
considered in Chapter \@ref(chapter2) to get more details on specific differences 
among *all* the pairs of groups -- what we call ***pair-wise comparisons***.
An issue is created when you perform many tests simultaneously and we will augment 
our previous methods with an adjusted method for pairwise comparisons to make our
results valid called ***Tukey's Honest Significant Difference***. 

\indent To make this more concrete, we return to the original MockJury data, making
side-by-side boxplots and beanplots (Figure \@ref(fig:Figure3-1)) as well as 
summarizing the suggested sentence lengths by the three groups using ``favstats``. 

```{r}
require(heplots)
require(mosaic)
data(MockJury)
favstats(Years~Attr, data=MockJury)
```

(ref:fig3-1) Boxplot and beanplot of the sentences (years) for the three 
treatment groups.

```{r Figure3-1,fig.cap="(ref:fig3-1)",echo=F,fig.height=5,fig.width=10}
require(heplots)
require(mosaic)
data(MockJury)
par(mfrow=c(1,2))
boxplot(Years~Attr,data=MockJury)
require(beanplot)
beanplot(Years~Attr,data=MockJury,log="",col="bisque",method="jitter")
```

```{r eval=F}
par(mfrow=c(1,2))
boxplot(Years~Attr,data=MockJury)
require(beanplot)
beanplot(Years~Attr,data=MockJury,log="",col="bisque",method="jitter")
```

There are slight differences in the sample sizes in the three groups with $37$
*Unattractive*, $38$ *Average* and $39$ *Beautiful* group responses, providing a
data set has a total sample size of $N=114$. The *Beautiful* and *Average* groups 
do not appear to be very different with means of 4.33 and 3.97 years. In 
Chapter \@ref(chapter2), we found moderate evidence regarding the difference in
*Average* and *Unattractive*. It is less clear whether we might find evidence of a
difference between *Beautiful* and *Unattractive* groups since we are comparing 
means of 5.81 and 4.33 years. All the distributions appear to be right skewed 
with relatively similar shapes. The variability in *Average* and *Unattractive* 
groups seems like it could be slightly different leading to an overall concern of
whether the variability is the same in all the groups. 

## Linear model for One-Way ANOVA (cell-means and reference-coding) {#section3-2}

We introduced the statistical model $y_{ij} = \mu_j+\varepsilon_{ij}$ in 
Chapter \@ref(chapter2) for the situation with $j = 1 \text{ or } 2$ to denote 
a situation where there were two groups and, for the model that is consistent 
with the alternative hypothesis, the means differed. Now we have three groups 
and the previous model can be extended to this new situation by allowing $j$
to be 1, 2, or 3. Now that we have more than two groups, we need to admit that 
what we were doing in Chapter \@ref(chapter2) was actually fitting what is called 
a ***linear model***.
\index{model!linear}
The linear model assumes that the responses follow a normal
distribution with the linear model defining the mean, all observations have the 
same variance, and the parameters for the mean in the model enter linearly. This 
last condition is hard to explain at this level of material -- it is sufficient 
to know that there are models where the parameters enter the model nonlinearly and 
that they are beyond the scope of this function and this material. By employing a general modeling methodology,  we will be able to use the same general modeling framework for the methods
in Chapters \@ref(chapter3), \@ref(chapter4), \@ref(chapter6),
\@ref(chapter7), and \@ref(chapter8). 

\indent As in Chapter \@ref(chapter2), we have a null hypothesis that defines a 
situation (and model) where all the groups have the same mean. Specifically, 
the ***null hypothesis*** in the general situation with $J$ groups 
($J\ge 2$) is to have all the $\underline{\text{true}}$ group means equal, 

$$H_0:\mu_1 = \ldots \mu_J.$$

This defines a model where all the groups have the same mean so it can be 
defined in terms of a single mean, $\mu$, for the $i^{th}$ observation from 
the $j^{th}$ group as $y_{ij} = \mu+\varepsilon_{ij}$. This is not the model 
that most researchers want to be the final description of their study as it 
implies no difference in the groups. There is more caution required to specify 
the alternative hypothesis with more than two groups. The 
***alternative hypothesis*** needs to be the logical negation of this null 
hypothesis of all groups having equal means; to make the null hypothesis
false, we only need one group to differ but more than one group could differ
from the others. Essentially, there are many ways to "violate" the null
hypothesis so we choose some delicate wording for the alternative hypothesis
when there are more than 2 groups. Specifically, we state the alternative as

$$H_A: \text{ Not all } \mu_j \text{ are equal}$$

or, in words, **at least one of the true means differs among the J groups**.
You will be attracted to trying to say that all means are different in the 
alternative but we do not put this strict a requirement in place to reject the 
null hypothesis. The alternative model
\index{model!alternative}
allows all the true group means to 
differ but does require that they differ with

$$y_{ij} = {\color{red}{\mu_j}}+\varepsilon_{ij}.$$

This linear model
\index{model!linear}
states that the response for the $i^{th}$ observation in 
the $j^{th}$ group, $\mathbf{y_{ij}}$, is modeled with a group $j$ 
($j=1, \ldots, J$) population mean, $\mu_j$, and a random error for each subject
in each group $\varepsilon_{ij}$, that we assume follows a normal distribution and 
that all the random errors have the same variance, $\sigma^2$. We can write the assumption about the random errors, often called the ***normality assumption***, 
as $\varepsilon_{ij} \sim N(0,\sigma^2)$. There is a second way to write out this 
model that allows extension to more complex models discussed below, so we
need a name for this version of the model. The model written in terms of the
${\color{red}{\mu_j}}\text{'s}$ is called the 
`r colFmtBold("cell means model","red")` and is the 
easier version of this model to understand. 
\index{model!cell-means}

\indent One of the reasons we learned about beanplots is that 
it helps us visually consider all the aspects of this model. 
\index{beanplot}
In the right panel of Figure \@ref(fig:Figure3-1),
we can see the wider, bold horizontal lines that provide the estimated group means.
The bigger the differences in the sample means, the more likely we are to find
evidence against the null hypothesis. You can also see the null model on the plot 
that assumes all the groups have the same mean as displayed in the
dashed horizontal line
at 4.7 years (the R code below shows the overall mean of *Years* is 4.7). While 
the hypotheses focus on the means, the model also contains assumptions about the
distribution of the responses -- specifically that the distributions are normal 
and that all the groups have the same variability.
\index{assumptions}
As discussed previously, it 
appears that the distributions are right skewed and the variability might not be 
the same for all the groups. The boxplot provides the information about the skew 
and variability but since it doesn't display the means it is not directly related 
to the linear model and hypotheses we are considering. 

```{r}
mean(MockJury$Years)
```

\indent There is a second way to write out the One-Way ANOVA model 
\index{model!One-Way ANOVA}
that provides a framework
for extensions to more complex models described in Chapter \@ref(chapter4) and 
beyond. The other ***parameterization*** (way of writing out or defining) of the 
model is called the `r colFmtBold("reference-coded model","purple")` since it 
writes out the model in terms of a 
\index{reference coding}
\index{model!reference-coded}
***baseline group*** and deviations from that baseline or reference level. The
reference-coded model for the $i^{th}$ subject in the $j^{th}$ group is 
$y_{ij} ={\color{purple}{\boldsymbol{\alpha + \tau_j}}}+\varepsilon_{ij}$ where 
$\color{purple}{\boldsymbol{\alpha}}$ (alpha) is the true mean for the
baseline group (first alphabetically) and the $\color{purple}{\boldsymbol{\tau_j}}$
(tau $j$) are the deviations from the baseline group for group $j$. The deviation 
for the baseline group, $\color{purple}{\boldsymbol{\tau_1}}$, is always set to 0 
so there are really just deviations for groups 2 through $J$. The equivalence 
between the two models can be seen by considering the mean for the first, second, 
and $J^{th}$ groups in both models:

$$\begin{array}{lccc}
& \textbf{Cell means:} && \textbf{Reference-coded:}\\
\textbf{Group } 1: & \color{red}{\mu_1} && \color{purple}{\boldsymbol{\alpha}} \\
\textbf{Group } 2: & \color{red}{\mu_2} && \color{purple}{\boldsymbol{\alpha + \tau_2}} \\
\ldots & \ldots && \ldots \\
\textbf{Group } J: & \color{red}{\mu_J} && \color{purple}{\boldsymbol{\alpha +\tau_J}}
\end{array}$$

The hypotheses for the reference-coded model are similar to those in the 
cell-means coding except that they are defined in terms of the deviations,
${\color{purple}{\boldsymbol{\tau_j}}}$. The null hypothesis is that there is
no deviation from the baseline for any group -- that all the ${\color{purple}{\boldsymbol{\tau_j\text{'s}}}}=0$,

$$\boldsymbol{H_0: \tau_2=\ldots=\tau_J=0}.$$

The alternative hypothesis is that at least one of the deviations is not 0, 

$$\boldsymbol{H_A:} \textbf{ Not all } \boldsymbol{\tau_j} \textbf{ equal } \bf{0}.$$

In this chapter, you are welcome to use either version (unless we instruct you
otherwise) but we have to use the reference-coding in subsequent chapters. The 
next task is to learn how to use R's linear model, ``lm``, function to get 
estimates of the parameters in each model, but first a quick review of these 
new ideas:

`r colFmtBold("Cell Means Version","red")`

* $H_0: {\color{red}{\mu_1=\ldots\mu_J}}$ &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp; $H_A: {\color{red}{\text{ Not all } \mu_j \text{ equal}}}$

* Null hypothesis in words: No difference in the true means between the groups. 

* Null model: $y_{ij} = \mu_j+\varepsilon_{ij}$ \index{model!null!cell-means}

* Alternative hypothesis in words: At least one of the true means differs between 
the groups. 

* Alternative model: $y_{ij} = \color{red}{\mu_j}+\varepsilon_{ij}.$
\index{model!alternative!cell-means}

`r colFmtBold("Reference-coded Version","purple")`

* $H_0: \color{purple}{\boldsymbol{\tau_2 \ldots \tau_J = 0}}$
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; 
$H_A: \color{purple}{\text{ Not all } \tau_j \text{ equal 0}}$

* Null hypothesis in words: No deviation of the true mean for any groups from the
baseline group. 

* Null model: $y_{ij} =\boldsymbol{\alpha} +\varepsilon_{ij}$
\index{model!null!reference-coded}

* Alternative hypothesis in words: At least one of the true deviations is 
different from 0 or that at least one group has a different true mean than the
baseline group. 

* Alternative model: $y_{ij} =\color{purple}{\boldsymbol{\alpha + \tau_j}}+\varepsilon_{ij}$ \index{model!alternative!reference-coded}

\indent In order to estimate the models discussed above, the ``lm`` function is used^[
If you look closely in the code for the rest of the book, any model for a 
quantitative response will use this function, suggesting a common thread in 
the most commonly used statistical models.]. The ``lm`` function continues to 
use the same format as previous functions, ``lm(Y~X, data=datasetname)``.
It ends up that this code will give you the reference-coded version of the 
model by default (The developers of R thought it was that important!).
\index{reference coding}
\index{\texttt{lm()}}
We want to start with the
cell-means version of the model,
\index{model!cell-means}
so we have to override the standard technique 
and add a "``-1``" to the formula interface to tell R that we want to the 
cell-means coding. Generally, this looks like ``lm(Y~X-1, data=datasetname).``
\index{\texttt{lm()}}
Once we fit a model in R, the ``summary`` function run on the model provides a
useful "summary" of the model coefficients and a suite of other potentially
interesting information. For the moment, we will focus on the estimated model
coefficients, so only those lines are output. When fitting this version
of the One-Way ANOVA model, 
you will find a row of output for each group relating the $\mu_j\text{'s}$.
The output contains columns for an estimate (``Estimate``), standard error 
(``Std. Error``), $t$-value (``t value``), and p-value (``Pr(>|t|)``). We'll 
learn to use all the output in the following material, but for now just focus 
on the estimates of the parameters that the function provides in the first
column ("Estimate") of the coefficient table. 

```{r}
lm1 <- lm(Years~Attr-1, data=MockJury)
summary(lm1)$coefficients
```

\indent In general, we denote estimated parameters with a hat over the parameter of 
interest to show that it is an estimate. For the true mean of group $j$, 
$\mu_j$, we estimate it with $\hat{\mu}_j$, which is just the sample mean for group
$j$, $\bar{x}_j$. The model suggests an estimate for each observation that we denote
as $\hat{y}_{ij}$ that we will also call a ***fitted value*** based on the model
being considered. The
same estimate is used for all observations in the each group. R tries to help you to 
sort out which row of output corresponds to which group by appending the group name
with the variable name. Here, the variable name was ``Attr`` and the first group
alphabetically was *Beautiful*, so R provides a row labeled ``AttrBeautiful``
with an estimate of 4.3333. The sample means from the three groups can be seen to
directly match for that group and the other two. 

```{r}
mean(Years~Attr, data=MockJury)
```

\indent The reference-coded version of the same model is more complicated but ends up 
\index{model!reference-coded}
giving the same results once we understand what it is doing. It uses a different
parameterization to accomplish this, so has different model output. Here is the model
summary:

```{r}
lm2 <- lm(Years~Attr, data=MockJury)
summary(lm2)$coefficients
```

The estimated model coefficients are $\hat{\alpha} = 4.333$ years, 
$\hat{\tau}_2 =-0.3596$ years, $\hat{\tau}_3=1.4775$ years where R selected group 1
for *Beautiful*, 2 for *Average*, and 3 for *Unattractive*. The way you can figure 
out the baseline group (group 1 is *Beautiful* here) is to see which category label 
is *not present* in the output. **The baseline level is typically the first group 
label alphabetically**, but you should always check this. Based on these definitions,
there are interpretations available for each coefficient. For $\hat{\alpha} = 4.333$ years, this is an estimate of the mean sentencing time for the *Beautiful* group.
$\hat{\tau}_2 =-0.3596$ years is the deviation of the *Average* group's mean from 
the *Beautiful* group's mean (specifically, it is $0.36$ years lower). Finally,
$\hat{\tau}_3=1.4775$ years tells us that the *Unattractive* group mean sentencing 
time is 1.48 years higher than the *Beautiful* group mean sentencing time. These
interpretations lead directly to
reconstructing the estimated means for each group by combining the baseline and
pertinent deviations as shown in Table \@ref(tab:Table3-1). 

(ref:tab3-1) Constructing group mean estimates from the reference-coded linear
model estimates.

```{r Table3-1, echo=F,results="asis"}
columnHeaders <- c("Group","Formula","Estimates")
row1 <- c("Beautiful","$\\hat{\\alpha}$","**4.3333** years")
row2 <- c("Average","$\\hat{\\alpha}+\\hat{\\tau}_2$", 
          "4.3333 - 0.3596 = **3.974** years")
row3 <- c("Unattractive","$\\hat{\\alpha}+\\hat{\\tau}_3$", 
          "4.3333 + 1.4775 = **5.811** years")
x <- matrix(c(row1,row2,row3),ncol=3,byrow=TRUE)
m <- as.data.frame(x)
colnames(m) <- columnHeaders
pandoc.table(m, keep.line.breaks = TRUE,caption="(\\#tab:Table3-1) (ref:tab3-1)",
             split.table = Inf, justify="left")
```

\indent We can also visualize the results of our linear models using what are called
***term-plots*** or ***effect-plots*** 
\index{effect-plot}
(from the ``effects`` package; [@R-effects])
\index{R packages!\textbf{effects}}
as displayed in Figure \@ref(fig:Figure3-2). We don't want to use the word 
"effect" for these model components unless we have random assignment in the study
\index{random assignment}
design so we generically call these ***term-plots*** as they display terms or
components from the model in hopefully useful ways to aid in model interpretation 
even in the presence of complicated model parameterizations. Specifically, these 
plots take an estimated model and show you its estimates along with 95% confidence
intervals generated by the linear model. To make this plot, you need to install and
load the ``effects`` package and then use ``plot(allEffects(...))`` functions 
together on the ``lm`` object called ``lm2`` that was estimated above. You can
find the correspondence between the displayed means and the estimates that were
constructed in Table \@ref(tab:Table3-1). 
\index{\texttt{allEffects()}}

(ref:fig3-2) Plot of the estimated group mean sentences from the reference-coded
model for the MockJury data from the ``effects`` package.

```{r Figure3-2, fig.cap="(ref:fig3-2)", fig.width=4,fig.height=2.75}
require(effects)
plot(allEffects(lm2))
```

\indent In order to assess evidence for having different means for the groups, we will 
compare either of the previous models (cell-means or reference-coded) to a null 
model based on the null hypothesis ($H_0: \mu_1 = \ldots = \mu_J$) which implies a
model of $\color{red}{y_{ij} = \mu_j}+\varepsilon_{ij}$ in the cell-means version 
where ${\color{red}{\mu}}$ is a common mean for all the observations. We will call 
this the `r colFmtBold("mean-only","red")` model since it only has a single mean 
in it. In the reference-coded version of the model, we have a null hypothesis that
$H_0: \tau_2 = \ldots = \tau_J = 0$, so the "mean-only" model is 
\index{model!mean-only}
$\color{purple}{y_{ij} =\boldsymbol{\alpha}+\varepsilon_{ij}}$ with 
$\color{purple}{\boldsymbol{\alpha}}$ having the same definition as 
$\color{red}{\mu}$ for the cell means model -- it forces a common value for the 
mean for all the groups. Moving from the *reference-coded* model to the *mean-only*
model is also an example of a situation where we move from a "full" model
\index{cell means}
\index{model!full}
to a
"reduced" model by setting some coefficients in the "full" model to 0 and, by doing
this, get a simpler or "reduced" model.
\index{model!reduced}
Simple models can be good as they are easier
to interpret, but having a model for $J$ groups that suggests no difference in the
groups is not a very exciting result
in most, but not all, situations^[Suppose we were doing environmental monitoring
and were studying asbestos levels in soils. We might be hoping that the mean-only 
model were reasonable to use if the groups being compared were in remediated areas 
and in areas known to have never been contaminated.]. In order for R to provide 
results for the mean-only model, we remove the grouping variable, ``Attr``, from 
the model formula and just include a "1". The ``(Intercept)`` row of the output
provides the estimate for the mean-only model as a reduced model from either the
cell-means or reference-coded models when we assume that the mean is the same
for all groups:

```{r}
lm3 <- lm(Years~1, data=MockJury)
summary(lm3)$coefficients
```

This model provides an estimate of the common mean for all observations of 
$4.693 = \hat{\mu} = \hat{\alpha}$ years. This value also is the dashed, horizontal
line in the beanplot in Figure \@ref(fig:Figure3-1). Some people 
call this mean-only model estimate the grand or overall mean. 

## One-Way ANOVA Sums of Squares, Mean Squares, and F-test {#section3-3}

The previous discussion showed two ways of parameterizing models for the 
One-Way ANOVA model and getting estimates from output but still hasn't 
addressed how to assess evidence related to whether the observed differences 
in the means among the groups is "real". In this section, we develop what is 
called the ***ANOVA F-test*** 
\index{F-test}
that provides a method of aggregating the 
differences among the means of 2 or more groups and testing our null hypothesis 
of no difference in the means vs the alternative. In order to develop the test, 
some additional notation is needed. The sample size in each group is denoted
$n_j$ and the total sample size is 
$\boldsymbol{N=\Sigma n_j = n_1 + n_2 + \ldots + n_J}$ where $\Sigma$
(capital sigma) means "add up over whatever follows". An estimated 
***residual*** ($e_{ij}$) is the difference between an observation, $y_{ij}$,
and the model estimate, $\hat{y}_{ij} = \hat{\mu}_j$, for that observation,
$y_{ij}-\hat{y}_{ij} = e_{ij}$. It is basically what is left over that the mean
part of the model ($\hat{\mu}_{j}$) does not explain. It is also a window
into how "good" the model might be because it reflects what the model was unable to explain. 

\indent Consider the four different fake results for a situation with four groups ($J=4$)
displayed in Figure \@ref(fig:Figure3-3). Which of the different results shows
the most and least evidence of differences in the means? In trying to answer
this, think about both how different the means are (obviously important) and
how variable the results are around the mean. These situations were created to
have the same means in Scenarios 1 and 2 as well as matching means in Scenarios
3 and 4. The variability around the means matches by shading (lighter or
darker). In Scenarios 1 and 2, the differences in the means is smaller than in
the other two results. But Scenario 2 should provide more evidence of what
little difference in present than Scenario 1 because it has less variability
around the means. The best situation for finding group differences here is
Scenario 4 since it has the largest difference in the means and the least
variability around those means. Our test statistic somehow needs to allow a
comparison of the variability in the means to the overall variability to help
us get results that reflect that Scenario 4 has the strongest evidence of a
difference and Scenario 1 would have the least. 

(ref:fig3-3) Demonstration of different amounts of difference in means relative 
to variability. Scenarios have same means in rows and same variance around means 
in columns of plot. 

```{r Figure3-3, fig.cap="(ref:fig3-3)", echo=F,fig.width=6}
set.seed(98876)
e1 <- rnorm(40)
x1 <- rep(c("A","B","C","D"),10)
mean1 <- rep(c(5,6,4,4.5),10)
mean2 <- rep(c(13,7,10,4),10)

y1 <- 0.5*e1+mean1
y2 <- 2*e1+mean1

y3 <- 0.5*e1+mean2
y4 <- 2*e1+mean2

par(mfrow=c(2,2))
old.par <-  par(mai=c(0.5,0.3,0.3,0.3))
beanplot(y2~x1,ylab="",main="Scenario 1",ylim=c(0,20),method="jitter",col="orange")
beanplot(y1~x1,ylab="",main="Scenario 2",ylim=c(0,20),method="jitter",col="red")

beanplot(y4~x1,ylab="",main="Scenario 3",ylim=c(0,20),method="jitter",col="orange")
beanplot(y3~x1,ylab="",main="Scenario 4",ylim=c(0,20),method="jitter",col="red")
```

\indent The statistic that allows the comparison of relative amounts of variation is called
the ***ANOVA F-statistic***. It is developed using ***sums of squares***
\index{sums of squares}
which are measures of total variation like those that are used in the numerator of the
standard deviation ($\Sigma_1^N(y_i-\bar{y})^2$) that took all the observations, 
subtracted the mean, squared the differences, and then added up the results
over all the observations to generate a measure of total variability. With
multiple groups, we will focus on decomposing that total variability 
(***Total Sums of Squares***) into variability among the means (we'll call this
***Explanatory Variable*** $\mathbf{A}\textbf{'s}$ ***Sums of Squares***) and
variability in the residuals \index{residuals}
or errors (***Error Sums of Squares***). We define each of these quantities in
the One-Way ANOVA situation as follows:

* $\textbf{SS}_{\textbf{Total}} =$ Total Sums of Squares 
$= \Sigma^J_{j=1}\Sigma^{n_j}_{i=1}(y_{ij}-\bar{\bar{y}})^2$

    * This is the total variation in the responses around the overall or
    ***grand mean*** ($\bar{\bar{y}}$, the estimated mean for all the 
    observations and available from the mean-only model). 

    * By summing over all $n_j$ observations in each group, $\Sigma^{n_j}_{i=1}(\ )$,
    and then adding those results up across the groups, $\Sigma^J_{j=1}(\ )$,
    we accumulate the variation across all $N$ observations. 

    * Note: this is the residual variation if the null model is used, so there 
    is no further decomposition possible for that model. 

    * This is also equivalent to the numerator of the sample variance, 
    $\Sigma^{N}_{1}(y_{i}-\bar{y})^2$ which is what you get when you ignore 
    the information on the potential differences in the groups. 

* $\textbf{SS}_{\textbf{A}} =$ Explanatory Variable *A*'s Sums of Squares 
$=\Sigma^J_{j=1}\Sigma^{n_j}_{i=1}(\bar{y}_{j}-\bar{\bar{y}})^2 =\Sigma^J_{j=1}n_j(\bar{y}_{j}-\bar{\bar{y}})^2$

    * This is the variation in the group means around the grand mean based on 
    the explanatory variable $A$.
    
    * Also called sums of squares for the treatment, regression, or model. 

* $\textbf{SS}_\textbf{E} =$ Error (Residual) Sums of Squares 
$=\Sigma^J_{j=1}\Sigma^{n_j}_{i=1}(y_{ij}-\bar{y}_j)^2 
=\Sigma^J_{j=1}\Sigma^{n_j}_{i=1}(e_{ij})^2$

    * This is the variation in the responses around the group means. 

    * Also called the sums of squares for the residuals, with the second 
    version of the formula showing that it is just the squared residuals added 
    up across all the observations. 

The possibly surprising result given the mass of notation just presented is that 
the total sums of squares is **ALWAYS** equal to the sum of explanatory variable
$A\text{'s}$ sum of squares and the error sums of squares, 

$$\textbf{SS}_{\textbf{Total}} \mathbf{=} \textbf{SS}_\textbf{A} \mathbf{+} \textbf{SS}_\textbf{E}.$$

This result is called the ***sums of squares decomposition formula***.
\index{sums of squares}
The equality means that if the $\textbf{SS}_\textbf{A}$ goes up, then the
$\textbf{SS}_\textbf{E}$ must go down if $\textbf{SS}_{\textbf{Total}}$ remains 
the same. We use these results to build our test statistic and organize this information in 
what is called an ***ANOVA table***.
\index{ANOVA table}
The ANOVA table is generated using the 
``anova`` function applied to the reference-coded model, ``lm2``:
\index{\texttt{anova()}}
\index{model!reference-coded}

```{r}
lm2 <- lm(Years~Attr, data=MockJury)
anova(lm2)
```

Note that the ANOVA table has a row labelled ``Attr``, which contains information 
for the grouping variable (we'll generally refer to this as explanatory variable 
$A$ but here it is the picture group that was randomly assigned), and a row 
labeled ``Residuals``, which is synonymous with "Error". The Sums of Squares 
(SS) are available in the ``Sum Sq`` column. It doesn't show a row for "Total" but 
the $\textbf{SS}_{\textbf{Total}} \mathbf{=} \textbf{SS}_\textbf{A} \mathbf{+} \textbf{SS}_\textbf{E} = 1492.26$.

```{r}
70.94 + 1421.32
```

```{r echo=F}
#Permutation approaches to an F-test:
set.seed(4321)
orig_anova <- anova(lm(Years~Attr, data=MockJury))
orig_SSA <- round(orig_anova[1,2], 1)
orig_SSE <- round(orig_anova[2,2], 1)
orig_SST <- orig_SSA + orig_SSE
orig_pval <- round(orig_anova[1,5], 3)
MockJuryP1 <- MockJury
MockJuryP1$Attr <- shuffle(MockJuryP1$Attr)
MJP1_anova <- anova(lm(Years~Attr, data=MockJuryP1))
MJP1_SSA <- round(MJP1_anova[1,2], 1)
MJP1_SSE <- round(MJP1_anova[2,2], 1)
MJP1_SST <- MJP1_SSA + MJP1_SSE
MockJuryP2 <- MockJury
MockJuryP2$Attr <- shuffle(MockJuryP2$Attr)
MJP2_anova <- anova(lm(Years~Attr, data=MockJuryP2))
MJP2_SSA <- round(MJP2_anova[1,2], 1)
MJP2_SSE <- round(MJP2_anova[2,2], 1)
MJP2_SST <- MJP2_SSA + MJP2_SSE
MockJuryP3 <- MockJury
MockJuryP3$Attr <- shuffle(MockJuryP3$Attr)
MJP3_anova <- anova(lm(Years~Attr, data=MockJuryP3))
MJP3_SSA <- round(MJP3_anova[1,2], 1)
MJP3_SSE <- round(MJP3_anova[2,2], 1)
MJP3_SST <- MJP3_SSA + MJP3_SSE
minSSA <- min(MJP1_SSA, MJP2_SSA, MJP3_SSA)
maxSSA <- max(MJP1_SSA, MJP2_SSA, MJP3_SSA)
```

```{r Figure3-4, fig.cap="(ref:fig3-4)",echo=F,fig.width=6}
Tobs <- anova(lm(Years~Attr, data=MockJury))[1,2]
par(mfrow=c(2,2))
old.par <- par(mai=c(0.5,0.3,0.3,0.3))
require(gplots)
label_list = c(paste("SSA=",as.character(orig_SSA),sep=""),
               paste("SSE=",as.character(orig_SSE),sep=""),
               paste("SSTotal=",as.character(orig_SST),sep=""))
plotmeans(Years~Attr,data=MockJury,ylab="Years",mean.labels=T,
          digits=2,legends=c("Beautiful","Average","Unattractive"),
          barcol="red",col="black", main="(a) Real Data",ylim=c(3,7.2))
text(c(1.25,1.25,1.25),c(7,6.5,6), labels=label_list, col="blue",cex=0.75)

label_list = c(paste("SSA=",as.character(MJP1_SSA),sep=""),
               paste("SSE=",as.character(MJP1_SSE),sep=""),
               paste("SSTotal=",as.character(MJP1_SST),sep=""))
plotmeans(Years~Attr,data=MockJuryP1,main="(b) Permuted Data",
          ylab="Years",mean.labels=T,digits=2,
          legends=c("Beautiful","Average","Unattractive"),
          barcol="red",col="black",ylim=c(3,7.2))
text(c(1.25,1.25,1.25),c(7,6.5,6),labels=label_list, col="blue",cex=0.75)

label_list = c(paste("SSA=",as.character(MJP2_SSA),sep=""),
               paste("SSE=",as.character(MJP2_SSE),sep=""),
               paste("SSTotal=",as.character(MJP2_SST),sep=""))
plotmeans(Years~Attr,data=MockJuryP2,main="(c) Permuted Data",
          ylab="Years",mean.labels=T,digits=2,
          legends=c("Beautiful","Average","Unattractive"),
          barcol="red",col="black",ylim=c(3,7.2))
text(c(1.25,1.25,1.25),c(7,6.5,6),labels=label_list, col="blue",cex=0.75)

label_list = c(paste("SSA=",as.character(MJP3_SSA),sep=""),
               paste("SSE=",as.character(MJP3_SSE),sep=""),
               paste("SSTotal=",as.character(MJP3_SST),sep=""))
plotmeans(Years~Attr,data=MockJuryP3,main="(d) Permuted Data",
          ylab="Years",mean.labels=T,digits=2,
          legends=c("Beautiful","Average","Unattractive"),
          barcol="red",col="black",ylim=c(3,7.2))
text(c(1.25,1.25,1.25),c(7,6.5,6),labels=label_list, col="blue",cex=0.75)
```

\indent It may be easiest to understand the *sums of squares decomposition* by connecting 
it to our permutation ideas. 
\index{sums of squares!decomposition}
\index{permutation}
In a permutation situation, the total variation 
($SS_\text{Total}$) cannot change -- it is the same responses varying
around the grand mean. However, the amount of variation attributed to variation 
among the means and in the residuals can change if we change which observations go 
with which group. In Figure \@ref(fig:Figure3-4) (panel a), the means, sums of 
squares, and 95% confidence intervals for each mean are displayed for the three
treatment levels from the original prisoner rating data. Three permuted versions 
of the data set are summarized in panels (b), (c), and (d). The $\text{SS}_A$ is 70.9 
in the real data set and between `r minSSA` and `r maxSSA` in the permuted data sets.
If you had 
to pick among the plots for the one with the most evidence of a difference in the
means, you hopefully would pick panel (a). This visual "unusualness" suggests
that this observed result is unusual relative to the possibilities under
permutations, which are, again, the possibilities tied to having the null
hypothesis being true. But note that the differences here are not that great
between these three permuted data sets and the real one. It is likely that at
least some might have selected panel (d) as also looking like it shows
some evidence of differences (maybe not the most?) as it looks like it
shows some evidence differences. 

\indent One way to think about $\textbf{SS}_\textbf{A}$ is that it is a function that 
converts the variation in the group means into a single value. This makes it a
reasonable test statistic in a permutation testing context. 
\index{permutation!test}
By comparing the 
observed $\text{SS}_A =$ `r orig_SSA` to the permutation results of 
`r MJP1_SSA`, `r MJP2_SSA`, and `r MJP3_SSA` we see 
that the observed result is much more extreme than the three alternate versions. 
In contrast to our previous test statistics where positive and negative 
differences were possible, $\text{SS}_A$ is always positive with a value of 0 
corresponding to no variation in the means. The larger the $\text{SS}_A$, the more 
variation there is in the means. The permutation p-value for the alternative 
hypothesis of **some** (not of greater or less than!) difference in the true 
means of the groups will involve counting the number of permuted $SS_A^*$ results
that are as large or larger than what we observed. 
\index{p-value!permutation distribution}

(ref:fig3-4) Plot of means and 95% confidence intervals for the three groups 
for the real Mock Jury data (a) and three different permutations of the treatment labels 
to the same responses in (b), (c), and (d). Note that ``SSTotal`` is always the same 
but the different amounts of variation associated with the means (``SSA``) or the 
errors (``SSE``) changes in permutation.

\indent To do a permutation test,
\index{permutation!test}
we need to be able to calculate and extract the 
$\text{SS}_A$ value. In the ANOVA table, it is the second number in the first row;
we can use the bracket, ``[,]``, referencing to extract that 
number from the ANOVA table that ``anova`` produces with 
``anova(lm(Years~Attr, data=MockJury))[1, 2]``. We'll store the observed value
of $\text{SS}_A$ in ``Tobs``, reusing some ideas from Chapter \@ref(chapter2). 
\index{\texttt{anova()}}

```{r}
Tobs <- anova(lm(Years~Attr, data=MockJury))[1,2]; Tobs
```

The following code performs the permutations ``B=1,000`` times using the 
``shuffle`` function, builds up a vector of results in ``Tobs``, and then makes 
a plot of the resulting permutation distribution:

(ref:fig3-5) Histogram and density curve of permutation distribution of 
$\text{SS}_A$ with the observed value of $\text{SS}_A$ displayed as a bold, 
vertical line. The proportion of results that are as large or larger than the observed 
value of $\text{SS}_A$ provides an estimate of the p-value. 

```{r Figure3-5,fig.cap="(ref:fig3-5)",echo=F,fig.height=5,fig.width=10}
par(mfrow=c(1,2))
B <- 1000
Tstar <- matrix(NA,nrow=B)
for (b in (1:B)){
  Tstar[b] <- anova(lm(Years~shuffle(Attr),data=MockJury))[1,2]
  }
hist(Tstar,labels=T,ylim=c(0,550))
abline(v=Tobs,col="red",lwd=3)
plot(density(Tstar),main="Density curve of Tstar")
abline(v=Tobs,col="red",lwd=3)
```

```{r eval=F}
B <- 1000
Tstar <- matrix(NA, nrow=B)
for (b in (1:B)){
  Tstar[b] <- anova(lm(Years~shuffle(Attr), data=MockJury))[1,2]
  }
hist(Tstar, labels=T, ylim=c(0,550))
abline(v=Tobs, col="red", lwd=3)
plot(density(Tstar), main="Density curve of Tstar")
abline(v=Tobs, col="red", lwd=3)
```

The right-skewed distribution (Figure \@ref(fig:Figure3-5)) contains the 
distribution of $\text{SS}^*_A\text{'s}$ under permutations (where
all the groups are assumed to be equivalent under the null hypothesis). While
the observed result is larger than many of the $\text{SS}^*_A\text{'s}$, there are
also many permuted results that are much larger than observed. The proportion
of permuted results that exceed the observed value is found using ``pdata`` 
as before, except only for the area to the right of the observed result. We know 
that ``Tobs`` will always be positive so no absolute values are required here. 
\index{\texttt{pdata()}}

```{r}
pdata(Tstar, Tobs, lower.tail=F)
```

```{r echo=F}
x <- pdata(Tstar, Tobs, lower.tail=F)
```

This provides a permutation-based p-value of `r x` and suggests marginal evidence
against the null hypothesis of no difference in the true means. We would interpret 
this p-value as saying that there is a `r 100*x`% chance of getting a $\text{SS}_A$
as large or larger than we observed, given that the null hypothesis is true.
\index{p-value!interpretation of}

\indent It ends up that some nice parametric statistical results 
are available (if our assumptions are met) for the ratio of estimated variances,
the estimated variances are called ***Mean Squares***.
\index{parametric}
To turn sums of squares into mean square (variance) estimates, 
we divide the sums of squares by the amount of free information available. For
example, remember the typical variance estimator introductory statistics,
$\Sigma^N_1(y_i-\bar{y})^2/(N-1)$? Your instructor spent some time trying various
approaches to explaining why we have a denominator of $N-1$. The most useful for our
purposes moving forward is that we "lose" one piece of information to estimate
the mean and there are $N$ deviations around the single mean so we divide by 
$N-1$. The main point is that the sums of squares were divided by something and 
we got an estimator for the variance, here of the observations overall. 


\indent Now consider $\text{SS}_E = \Sigma^J_{j=1}\Sigma^{n_j}_{i=1}(y_{ij}-\bar{y}_j)^2$
which still has $N$ deviations but it varies around the $J$ means, so the

$$\textbf{Mean Square Error} = \text{MS}_E = \text{SS}_E/(N-J).$$

Basically, we lose $J$ pieces of information in this calculation because we have 
to estimate $J$ means. The similar calculation of the ***Mean Square for variable*** $\mathbf{A}$
($\text{MS}_A$) is harder to see in the formula 
($\text{SS}_A = \Sigma^J_{j=1}n_j(\bar{y}_i-\bar{\bar{y}})^2$), but the same 
reasoning can be used to understand the denominator for forming $\text{MS}_A$:
there are $J$ means that vary around the grand mean so

$$\text{MS}_A = \text{SS}_A/(J-1).$$

In summary, the two mean squares are simply:

* $\text{MS}_A = \text{SS}_A/(J-1)$, which estimates the variance of the group 
means around the grand mean. 

* $\text{MS}_{\text{Error}} = \text{SS}_{\text{Error}}/(N-J)$, which estimates 
the variation of the errors around the group means.

\newpage

\indent These results are put together using a ratio to define the ***ANOVA F-statistic***
(also called the ***F-ratio***) as: 

$$F=\text{MS}_A/\text{MS}_{\text{Error}}.$$

If the variability in the means is "similar" to the variability in the residuals,
the statistic would have a value around 1. If that variability is similar then 
there would be no evidence of a difference in the means. If the $\text{MS}_A$ is much 
larger than the $\text{MS}_E$, the $F$-statistic will provide evidence against 
the null hypothesis. The "size" of the $F$-statistic is formalized by finding the
p-value. The $F$-statistic, if assumptions discussed below are met and we assume 
the null hypothesis is true, follows what is called an $F$-distribution.
\index{F-distribution}
The 
***F-distribution*** is a right-skewed distribution whose shape is defined by what
are called the ***numerator degrees of freedom*** ($J-1$) and the
***denominator degrees of freedom*** ($N-J$). These names correspond to the values
that we used to calculate the mean squares and where in the $F$-ratio each mean 
square was used; $F$-distributions are denoted by their degrees of freedom using 
the convention of $F$ (*numerator df*, *denominator df*). Some examples of 
different $F$-distributions are displayed for you in Figure \@ref(fig:Figure3-6).

\indent The characteristics of the F-distribution can be summarized as:

* Right skewed,

* Nonzero probabilities for values greater than 0, 

* Its shape changes depending on the **numerator** and **denominator DF**, and

* **Always use the right-tailed area for p-values.**


(ref:fig3-6) Density curves of four different $F$-distributions. Upper left is an 
$F(2, 111)$, upper right is $F(2, 10)$, lower left is $F(6, 10)$, and lower right 
is $F(6, 111)$. P-values are found using the areas to the right of the observed
$F$-statistic value in all F-distributions. \index{p-value!calculation of}

```{r Figure3-6,fig.cap="(ref:fig3-6)",echo=F,fig.height=4.25}
x <- seq(from=0,to=6, length.out=50)
par(mfrow=c(2,2))
old.par <- par(mai=c(0.5,0.3,0.3,0.3))
plot(x,df(x,df1=2,df2=111),type="l",ylab="Density",xlab="F",
     main="F(2,111)",col="red",lwd=3)

plot(x,df(x,df1=2,df2=10),type="l",ylab="Density",xlab="F",
     main="F(2,10)",col="blue",lwd=3)

plot(x,df(x,df1=6,df2=10),type="l",ylab="Density",xlab="F",
     main="F(6,10)",col="green",lwd=3)

plot(x,df(x,df1=6,df2=111),type="l",ylab="Density",xlab="F",
     main="F(6,111)",col="magenta",lwd=3)
```

\indent Now we are ready to discuss an ANOVA table since we know about each of its 
components. Note the general format of the ANOVA table is in Table \@ref(tab:Table3-2)^[Make sure you can work 
from left to right and up and down to fill in the ANOVA table given just the 
necessary information to determine the other components -- there is always a 
question like this on the exam...]:
\index{ANOVA table}

(ref:tab3-2) General One-Way ANOVA table.

```{r Table3-2, echo=F, message=F,warning=F,results="asis"}
columnHeaders <- c("Source&nbsp;&nbsp;","DF&nbsp;","Sums of\\\n Squares",
                   "Mean Squares","F-ratio","P-value")
row1 <- c("Variable A","$J-1$","$\\text{SS}_A$",
          "$\\text{MS}_A=\\text{SS}_A/(J-1)$",
          "$F=\\text{MS}_A/\\text{MS}_E$","Right tail of $F(J-1,N-J)$")
row2 <- c("Residuals","$N-J$","$\\text{SS}_E$","$\\text{MS}_E = \\text{SS}_E/(N-J)$","","")
row3 <- c("Total","$N-1$","$\\text{SS}_{\\text{Total}}$","","","")
x <- matrix(c(row1,row2,row3),ncol=6,byrow=TRUE)
m <- as.data.frame(x)
colnames(x) <- columnHeaders
pandoc.table(x, keep.line.breaks = TRUE,caption="(\\#tab:Table3-2) (ref:tab3-2)",
             split.table = Inf, justify="left")
```

The table is oriented to help you reconstruct the $F$-ratio from each of its
components. The output from R is similar although it does not provide the last row 
and sometimes switches the order of columns in different functions we will use. The R version of the table for the type 
of picture effect (``Attr``) with $J=3$ levels and $N=114$ observations, repeated 
from above, is:

```{r}
anova(lm2)
```

The p-value from the $F$-distribution is `r round(orig_anova[1,5],3)`. We can 
verify this result using the observed $F$-statistic of `r round(orig_anova[1,4],2)`
(which came from taking the ratio of the two mean squares, 
F=`r round(orig_anova[1,3],2)`/`r round(orig_anova[2,3],2)`)
which follows an $F(2, 111)$ distribution if the null hypothesis is true and some 
other assumptions are met. 

Using the ``pf`` function provides us with areas in the
specified $F$-distribution with the ``df1`` provided to the function as the 
numerator *df* and ``df2`` as the denominator *df* and ``lower.tail=F`` reflecting 
our desire for a right tailed area.
\index{\texttt{pf()}}

```{r}
pf(2.77, df1=2, df2=111, lower.tail=F)
```

\indent The result from the $F$-distribution using this parametric procedure is similar to 
the p-value obtained using permutations with the test statistic of the 
$\text{SS}_A$, which was `r orig_pval`. The $F$-statistic obviously is another
potential test statistic to use as a test statistic in a permutation approach, 
now that we know about it. We should check that we get similar results from it
with permutations as we did from using $\text{SS}_A$ as a permutation-test test
statistic. The following code generates the permutation distribution
\index{permutation!distribution}
for the
$F$-statistic (Figure \@ref(fig:Figure3-7)) and assesses how unusual the observed
$F$-statistic of `r round(orig_anova[1,4],2)` was in this permutation distribution.
The only change in the code involves moving from extracting $\text{SS}_A$ to 
extracting the $F$-ratio which is in the 4^th^ column of the ``anova`` 
output:

```{r}
Tobs <- anova(lm(Years~Attr, data=MockJury))[1,4]; Tobs
B <- 1000
Tstar <- matrix(NA, nrow=B)
for (b in (1:B)){
  Tstar[b] <- anova(lm(Years~shuffle(Attr), data=MockJury))[1,4]
}

pdata(Tstar, Tobs, lower.tail=F)
```

```{r eval=F}
hist(Tstar, labels=T)
abline(v=Tobs, col="red", lwd=3)
plot(density(Tstar), main="Density curve of Tstar")
abline(v=Tobs, col="red", lwd=3)
```

(ref:fig3-7) Histogram and density curve of the permutation distribution of 
the F-statistic with bold, vertical line for the observed value of the test 
statistic of `r round(orig_anova[1,4],2)`. 

```{r Figure3-7,fig.cap="(ref:fig3-7)",echo=F,fig.height=5,fig.width=10}
par(mfrow=c(1,2))
hist(Tstar, labels=T, ylim=c(0,550))
abline(v=Tobs, col="red", lwd=3)
plot(density(Tstar), main="Density curve of Tstar")
abline(v=Tobs, col="red", lwd=3)
```

```{r echo=F}
perm_pval <- pdata(Tstar, Tobs, lower.tail=F)
```

The permutation-based p-value is `r perm_pval` which, again, matches the other 
results closely. The first conclusion is that using a test statistic of either 
the $F$-statistic or the $\text{SS}_A$ provide similar permutation results.
However, we tend to favor using the $F$-statistic because it is more commonly used 
in reporting ANOVA results, not because it is any better in a permutation context. 


\indent It is also interesting to compare the permutation distribution for the 
$F$-statistic and the parametric $F(2, 111)$ distribution 
(Figure \@ref(fig:Figure3-8)). They do not match perfectly but are quite similar. 
Some the differences around 0 are due to the behavior of the method used to create 
the density curve and are not really a problem for the methods. The similarity in
the two curves explains why both methods give similar results. In some
situations, the correspondence will not be quite so close. 

(ref:fig3-8) Comparison of $F(2, 111)$ (dashed line) and permutation distribution
(solid line). 

```{r Figure3-8,fig.cap="(ref:fig3-8)",echo=F,fig.height=3}
par(mfrow=c(1,1))
plot(density(Tstar), xlab="F-values", ylab="Density",
     main="Comparison of permutation and F(2,111) distributions",
     lwd=2, xlim=c(0,9), cex.main=0.75)
x <- seq(from=0,to=9, length.out=50)
lines(x, df(x,df1=2,df2=111), col="red", lwd=2, lty=2)
legend("right", legend=c("Permutation distribution","F(2,111)"), col=c("black","red"), lty=c(1,2), lwd=2, cex=0.5)
```

\indent So how can we rectify this result ($\text{p-value}\approx 0.06$) and the 
Chapter \@ref(chapter2) result that detected a difference between *Average* 
and *Unattractive* with a $\text{p-value}\approx 0.03$? I selected the two groups 
to compare in Chapter \@ref(chapter2) because they were furthest apart.
"Cherry-picking" the comparison that is likely to be most different creates a 
false sense of the real situation and inflates the Type I error rate because of 
the selection^[This fits with a critique of p-value usage called 
p-hacking or publication bias -- where researchers search across many 
results and only report their biggest differences. This biases the 
results to detecting results more than they should be and then when 
other researchers try to repeat the same studies, they fail to find 
similar results.]. 
\index{p-value!criticisim}
\index{type I error}
If the entire suite of pairwise comparisons are considered, this 
result may lose some of its luster. In other words, if we consider the suite of 
three pair-wise differences (and the tests) implicit in comparing all of them, 
we may need stronger evidence in the most different pair than a 
p-value of 0.033 to suggest overall differences. In this situation, 
the *Beautiful* and *Average* 
groups are not that different from each other so their difference does not 
contribute much to the overall $F$-test. In Section \@ref(section3-6), we will 
revisit this topic and consider a method that is
statistically valid for performing all possible pair-wise comparisons that is also
consistent with our overall test results. 

## ANOVA model diagnostics including QQ-plots {#section3-4}

The requirements for a One-Way ANOVA $F$-test are similar to those discussed in 
Chapter \@ref(chapter2), except that there are now $J$ groups instead of only 2.
Specifically, the linear model assumes:

1. **Independent observations**,

2. **Equal variances**, and

3. **Normal distributions**. 

\indent For assessing equal variances across the groups, it is best to use plots to  assess this. We can use boxplots and beanplots to compare the spreads of the 
groups, which were provided in Figure \@ref(fig:Figure3-1). The range and IQRs 
should be relatively similar across the groups if you do not find evidence of a 
problem with this assumption. You should start with noting how clear or big the
violation of the assumption might be but remember that there will always be some
differences in the variation among groups even if the true variability is exactly 
equal in the populations. In addition to our direct plotting, there are some 
diagnostic plots available from the ``lm`` function that can help us more
clearly assess potential violations of the previous assumptions.
\index{\texttt{lm()}}

\indent We can obtain a suite of four diagnostic plots by using the ``plot`` function on 
any linear model object that we have fit. To get all the plots together in four 
panels we need to add the ``par(mfrow=c(2,2))`` command to tell R to make a graph
with 4 panels^[We have been using this function quite a bit to make multi-panel 
graphs but did not show you that line of code. But you need to use this command 
for linear model diagnostics or you won't get the plots we want from the model. 
And you really just need ``plot(lm2)`` but the ``pch=16`` option makes it easier 
to see some of the points in the plots.].

```{r eval=F}
par(mfrow=c(2,2))
plot(lm2, pch=16)
```

There are two plots in Figure \@ref(fig:Figure3-9) with useful information for assessing the
equal variance assumption. The "Residuals vs Fitted" panel in the top left panel displays  the residuals $(e_{ij} = y_{ij}-\hat{y}_{ij})$ on the y-axis and the fitted values
$(\hat{y}_{ij})$ on the x-axis. 
\index{Residuals vs Fitted plot}
This allows you to see if the variability of the
observations differs across the groups as a function of the mean of the groups, 
because all the observations in the same group get the same fitted value -- the 
mean of the group. In this plot, the points seem to have fairly similar spreads 
at the fitted values for the three groups with fitted values of 4, 4.3, and 6. 
The "Scale-Location" plot in the lower left panel has the same x-axis but the 
y-axis contains the square-root of the absolute value of the standardized 
residuals.
\index{Scale-Location plot}
The absolute value transforms all the residuals into a magnitude 
scale (removing direction) and the square-root helps you see differences in
variability more accurately. The standardization scales the residuals to have a variance 
of 1 so help you in other displays to get a sense of how many standard deviations 
you are away from the mean in the residual distribution. The visual assessment is
similar in the two plots -- you want to consider whether it appears that the 
groups have somewhat similar or noticeably different amounts of variability. If 
you see a clear funnel shape in the Residuals vs Fitted 
\index{Residuals vs Fitted!interpretation of}
or an increase or decrease 
in the upper edge of points in the Scale-Location plot that may indicate a 
violation of the constant variance assumption.
\index{Scale-Location plot!interpretation of}
Remember that some variation 
across the groups is expected and is OK, but large differences in spreads are problematic for all the procedures that involve linear models. When discussing 
these results, you want to discuss how clearly the differences in variation are 
and whether that *shows a clear violation of the assumption* of equal variance 
for all observations. Like in hypothesis testing, you can never prove that you've 
met assumptions based on a plot "looking OK", but you can say that there is no 
clear evidence that the assumption is violated! 
\index{assumptions}

(ref:fig3-9) Default diagnostic plots for the Mock Jury full linear model.

```{r Figure3-9,fig.cap="(ref:fig3-9)",echo=F,fig.width=10,fig.height=5}
par(mfrow=c(2,2))
plot(lm2, pch=16)
```

\indent The linear model also assumes that all the random errors ($\varepsilon_{ij}$) follow a 
normal distribution. To gain insight into the validity of this assumption, we 
can explore the original observations as displayed in the beanplots, mentally
subtracting off the differences in the means and focusing on the shapes of the
distributions of observations in each group. These plots are especially good for
assessing whether there is a skew or outliers present in each group. If so, 
by definition, the normality assumption is violated. But our assumption is 
about the distribution of all the errors after removing the differences 
in the means and so we want an overall assessment technique to understand how
reasonable our assumption is overall for our model. The residuals from the entire
\index{residuals}
model provide us with estimates of the random errors and if the normality 
assumption is met, then the residuals all-together should approximately follow a
normal distribution. The ***Normal Q-Q Plot*** in the upper right panel of 
Figure \@ref(fig:Figure3-9) is a direct visual assessment of how well our 
residuals match what we would expect from a normal distribution. Outliers, skew, 
heavy and light-tailed aspects of distributions (all violations of normality) 
show up in this plot once you learn to read it -- which is our next task. To 
make it easier to read QQ-plots, it is nice to start with just considering 
histograms and/or density plots of the residuals and to see how that maps into 
this new display. We can obtain the residuals from the linear model using the 
``residuals`` function on any linear model object. 

(ref:fig3-10) Histogram and density curve of the linear model raw residuals from the Mock Jury linear model. 

```{r Figure3-10, fig.cap="(ref:fig3-10)",fig.height=5,fig.width=10}
par(mfrow=c(1,2))
eij <- residuals(lm2)
hist(eij, main="Histogram of residuals")
plot(density(eij), main="Density plot of residuals", ylab="Density",
     xlab="Residuals")
```

Figure \@ref(fig:Figure3-10) shows that there is a right skew present in the 
residuals for the model for the prisoner ratings that accounted for different 
means in the three picture groups, which is consistent with the initial assessment of 
some right skew in the plots of observations in each group.

\indent A Quantile-Quantile plot (***QQ-plot***) 
\index{QQ-plot}
shows the "match" of an observed 
distribution with a theoretical distribution, almost always the normal
distribution. They are also known as Quantile Comparison, Normal Probability, 
or Normal Q-Q plots, with the last two names being specific to comparing
results to a normal distribution. In this version^[Along with multiple names,
there is variation of what is plotted on the x and y axes and the scaling of 
the values plotted, increasing the challenge of interpreting QQ-plots. We are
consistent about the x and y axis choices throughout this book but different functions that make 
these plots in R do switch the axes.], the QQ-plots display the value of 
observed percentiles in the residual distribution on the y-axis versus the 
percentiles of a theoretical normal distribution on the x-axis. If the 
observed **distribution of the residuals matches the shape of the normal
distribution, then the plotted points should follow a 1-1 relationship.**
If the points follow the displayed straight line then that suggests that the 
residuals have a similar shape to a normal distribution. Some variation is 
expected around the line and some patterns of deviation are worse
than others for our models, so you need to go beyond saying "it does not match
a normal distribution". It is best to be specific about the type of deviation
you are detecting. And to do that, we need to practice interpreting some
QQ-plots. 

\indent The QQ-plot of the linear model residuals from Figure \@ref(fig:Figure3-9) is extracted and enhanced it a little to make Figure \@ref(fig:Figure3-11) so we 
can just focus on it.
\index{QQ-plot!interpretation of}
We know from looking at the histogram that this is a 
slightly right skewed distribution. The QQ-plot places the observed ***standardized***^[Here this means re-scaled so that they should have similar 
scaling to a standard normal with mean 0 and standard deviation 1. This does 
not change the shape of the distribution but can make outlier identification 
simpler -- having a standardized residual more extreme than 5 or -5 would 
suggest a deviation from normality since we rarely see values that many 
standard deviations from the mean in a normal distribution. But mainly focus 
on the shape of the pattern in the QQ-plot.] ***residuals*** on the y-axis 
and the theoretical normal values on the x-axis. The most noticeable deviation 
from the 1-1 line is in the lower left corner of the plot. These are for the 
negative residuals (left tail) and there are many residuals at around the same 
value that are a little smaller than -1. If the distribution had followed the 
normal distribution here, the points would be on the 1-1 line and there would 
be some standardized residuals much smaller than -1.5. So we are not getting as 
much spread in the smaller residuals as we would expect in a normal distribution.
If you go back to the histogram you can see that the smallest residuals are all
stacked up and do not spread out like the left tail of a normal distribution 
should. In the right tail (positive) residuals, there is also a systematic 
lifting from the 1-1 line to larger values in the residuals than the normal 
would generate. For example, the point labeled as "82" in Figure \@ref(fig:Figure3-9)  (the 82^nd^
observation in the data set) has a value of 3 in residuals but should actually 
be smaller (maybe 2.5) if the distribution was normal. Put together, this pattern 
in the QQ-plot suggests that the left tail is too compacted (too short) and 
the right tail is too spread out -- this is the right skew we identified from the
histogram and density curve! 

(ref:fig3-11) QQ-plot of residuals from Mock Jury linear model.


```{r Figure3-11,fig.cap="(ref:fig3-11)",echo=F,fig.height=3,fig.width=6}
par(mfrow=c(1,2))
require(car)
qqPlot(eij, main="QQ-Plot of residuals")
plot(density(eij), main="Density plot of residuals", ylab="Density",
     xlab="Residuals")
```


\indent Generally, when both tails deviate on the same side of the line (forming a 
sort of quadratic curve, especially in more extreme cases), that is evidence 
of a skew. To see some different potential shapes in QQ-plots, six different 
data sets are displayed in Figures \@ref(fig:Figure3-12) and \@ref(fig:Figure3-13).
In each row, a QQ-plot and associated density curve are displayed. If the points 
are both above the 1-1 line in the lower and upper tails as in 
Figure \@ref(fig:Figure3-12)(a), then the pattern is a right skew, here even more
extreme than in the previous real data set. If the points are below the 1-1 line in both
tails as in Figure \@ref(fig:Figure3-12)(c), then the pattern is identified as a 
left skew. Skewed residual distributions (either direction) are problematic for 
models that assume normally distributed responses but not necessarily for our
permutation approaches if all the groups have similar skewed shapes. The other
problematic pattern is to have more spread than a normal curve as in 
Figure \@ref(fig:Figure3-12)(e) and (f). This shows up with the points being 
below the line in the left tail (more extreme negative than expected by the normal)
and the points being above the line for the right tail (more extreme positive 
than the normal predicts). We call these distributions ***heavy-tailed***
which can manifest as distributions with outliers in both tails or just a bit 
more spread out than a normal distribution. Heavy-tailed residual distributions 
can be problematic for our models as the variation is greater than what the normal
distribution can account for and our methods might under-estimate the 
variability in the results. The opposite pattern with the left tail above the 
line and the right tail below the line suggests less spread (***lighter-tailed***)
than a normal as in Figure \@ref(fig:Figure3-12)(g) and (h). This pattern is
relatively harmless and you can proceed with methods that assume normality safely 
as they will just be a little conservative. For any of the patterns, you would 
note a potential violation of the normality assumption and then proceed to 
describe the type of violation and how clear or extreme it seems to be.

(ref:fig3-12) QQ-plots and density curves of four simulated distributions with
different shapes. 

```{r Figure3-12,fig.cap="(ref:fig3-12)",echo=F,fig.height=8,fig.width=6}
#QQ plot code and examples:
require(car)
par(mfrow=c(4,2))
#Right skew
y1 <- scale(rf(75,df1=3,df2=8))
qqPlot(y1,main="(a) QQ-Plot of y1",envelope=F,pch=19,cex=1.3)
plot(density(y1),xlab="y1",ylab='Density',main="(b) Density plot of y1",col="blue",xlim=c(-6.5,6.5))
#Left Skew
y2 <- scale(-rf(75,df1=3,df2=8))
qqPlot(y2,main="(c) QQ-Plot of y2",envelope=F,pch=19,cex=1.3)
plot(density(y2),xlab="y2",ylab='Density',main="(d) Density plot of y2",col="blue",xlim=c(-6.5,6.5))

#t(2)
y3 <- scale(rt(75,df=4))
qqPlot(y3,main="(e) QQ-Plot of y3",envelope=F,pch=19,cex=1.3)
plot(density(y3),xlab="y3",ylab='Density',main="(f) Density plot of y3",col="blue",xlim=c(-6.5,6.5))

#Uniform
y4 <- scale(runif(75,0,1))
qqPlot(y4,main="(g) QQ-Plot of y4",envelope=F,pch=19,cex=1.3)
plot(density(y4),xlab="y4",ylab='Density',main="(h) Density plot of y4",col="blue",xlim=c(-6.5,6.5))
```

\indent Finally, to help you calibrate expectations for data that are actually normally
distributed, two data sets simulated from normal distributions are displayed in 
Figure \@ref(fig:Figure3-13). Note how neither follows the line exactly but 
that the overall pattern matches fairly well. **You have to allow for some
variation from the line in real data sets** and focus on when there are really
noticeable issues in the distribution of the residuals such as those
displayed above. Again, you will never be able to prove that you have normally
distributed residuals even if the residuals are all exactly on the line, but if
you see QQ-plots as in Figure \@ref(fig:Figure3-12) you can determine that there is clear evidence of violations of the normality assumption. 
\index{QQ-plot!interpretation of}

(ref:fig3-13) Two more simulated data sets, both generated from normal distributions.

```{r Figure3-13,fig.cap="(ref:fig3-13)",echo=F,fig.width=6}
par(mfrow=c(2,2))
old.par <- par(mai=c(0.75,0.5,0.5,0.5))
#One normal example

y5 <- scale(rnorm(75))
qqPlot(y5,main="(i) QQ-Plot of y5",envelope=F,pch=19,cex=1.3)
plot(density(y5),xlab="y5",ylab='Density',main="(j) Density plot of y5",col="blue",xlim=c(-6.5,6.5))


#Another normal example
y6 <- scale(rnorm(75))
qqPlot(y6,main="(k) QQ-Plot of y6",envelope=F,pch=19,cex=1.3)
plot(density(y6),xlab="y6",ylab='Density',main="(l) Density plot of y6",col="blue",xlim=c(-6.5,6.5))
```

\indent The last issues with assessing the assumptions in an ANOVA relates to 
situations where the methods are more or less ***resistant***^[A resistant 
procedure is one that is not severely impacted by a particular violation of an
assumption. For example, the median is resistant to the impact of
an outlier. But the mean is not a resistant measure as changing the value
of a single point changes the mean.] to violations of assumptions. 
\index{resistant}
In simulation studies of the performance of the $F$-test, researchers 
have found that the
parametric ANOVA $F$-test is more resistant to violations of the assumptions of
the normality and equal variance assumptions if the design is balanced. 
\index{balanced}
A ***balanced design*** occurs when each group is measured the same number of 
times. The resistance decreases as the data set becomes less balanced, as the 
sample sizes in the groups are more different, so having close to balance is 
preferred to a more imbalanced situation if there is a choice available. There 
is some intuition available here -- it makes some sense that you would have better
results in comparing groups if the information available is similar in all the
groups and none are relatively under-represented. We can check the number of
observations in each group to see if they are equal or similar using the 
``tally`` function from the ``mosaic`` package. This function is useful for
being able to get counts of observations, especially for cross-classifying
observations on two variables that is used in Chapter \@ref(chapter5). For just 
a single variable, we use ``tally(~x, data=...)``:
\index{\texttt{tally()}}

```{r,warning=F,message=F}
require(mosaic)
tally(~Attr, data=MockJury)
```

So the sample sizes do vary among the groups and the design is technically not
balanced, but it is also very close to being balanced with only two more 
observations in the largest group compared to the smallest group size. This 
tells us that the $F$-test should have some resistance to violations of 
assumptions. This nearly balanced design, and the moderate sample size 
(over 37 per group is considered a good but not large sample), make the 
parametric and nonparametric approaches provide similar results in this data 
set even in the presence of the skewed residual error distribution that presents a violation to the assumptions of the parametric procedure. 



## Guinea pig tooth growth One-Way ANOVA example {#section3-5}

A second example of the One-way ANOVA methods involves a study of length of
odontoblasts (cells that are responsible for tooth growth) in 60 Guinea Pigs 
(measured in microns) from @Crampton1947. $N=60$ Guinea Pigs were obtained 
from a local breeder and each received one of three dosages (0.5, 1, or 
2 mg/day) of Vitamin C via one of two delivery methods, Orange Juice (*OJ*) or
ascorbic acid (the stuff in vitamin C capsules, called $\text{VC}$ below) as the source 
of Vitamin C in their diets. Each guinea pig was randomly assigned to receive 
one of the six different treatment combinations possible
(OJ at 0.5 mg, OJ at 1 mg, OJ at 2 mg, VC at 0.5 mg, VC at 1 mg, and VC at 2 mg). 
The animals were treated similarly otherwise and we can assume lived in
separate cages and only one observation was taken for each guinea pig, so we
can assume the observations are independent^[A violation of the independence
assumption could have easily been created if they measured cells in 
\index{independence assumption!One-Way ANOVA}
two locations on each guinea pig or took measurements over time on each subject.]. We need to create a variable that
combines the levels of delivery type (OJ, VC) and the dosages (0.5, 1, and 2)
to use our One-Way ANOVA on the six levels. The ``interaction`` function can be 
used create a new variable that is based on combinations of the levels of other
variables. Here a new variable is created in the ``ToothGrowth`` tibble that 
we called ``Treat`` using the ``interaction`` function that provides a six-level grouping variable for our One-Way 
ANOVA to compare the combinations of treatments. To get a sense of the pattern of
observations in the data set, the counts in ``supp`` (supplement type) and 
``dose`` are provided and then the counts in the new categorical explanatory variable, ``Treat``. 

```{r}
data(ToothGrowth) #Available in Base R
require(tibble)
ToothGrowth <- as.tibble(ToothGrowth) #Convert data.frame to tibble
require(mosaic)
tally(~supp, data=ToothGrowth) #Supplement Type (VC or OJ)
tally(~dose, data=ToothGrowth) #Dosage level

#Creates a new variable Treat with 6 levels
ToothGrowth$Treat <- with(ToothGrowth, interaction(supp, dose)) 

#New variable that combines supplement type and dosage
tally(~Treat, data=ToothGrowth) 
```

The ``tally`` function helps us to check for balance;
\index{balanced}
this is a balanced design
because the same number of guinea pigs ($n_j=10 \text{ for } j=1, 2,\ldots, 6$) 
were measured in each treatment combination. 

\indent With the variable ``Treat`` prepared, the first task is to visualize the results 
using boxplots and beanplots^[Note that to see all the group labels in the plot 
when making the figure, you have to widen the plot window before copying the 
figure out of R. You can resize the plot window using the small vertical and
horizontal "=" signs in the grey bars that separate the different panels in 
RStudio.] (Figure \@ref(fig:Figure3-14)) and generate some summary statistics for 
each group using ``favstats``. 

```{r}
favstats(len~Treat, data=ToothGrowth)
```

(ref:fig3-14) Boxplot and beanplot of odontoblast growth responses for the six 
treatment level combinations. 

```{r Figure3-14, fig.cap="(ref:fig3-14)",fig.height=6,fig.width=12}
par(mfrow=c(1,2))
boxplot(len~Treat, data=ToothGrowth, ylab="Odontoblast Growth in microns")
beanplot(len~Treat, data=ToothGrowth, log="", col="yellow",
         method="jitter", ylab="Odontoblast Growth in microns")
```

\indent Figure \@ref(fig:Figure3-14) suggests that the mean tooth growth increases with 
the dosage level and that OJ might lead to higher growth rates than VC except 
at a dosage of 2 mg/day. The variability around the means looks to be small 
relative to the differences among the means, so we should expect a small 
p-value from our $F$-test. The design is balanced as noted above ($n_j=10$ 
for all six groups) so the methods are some what resistant to impacts from
non-normality and non-constant variance but we should still 
assess the patterns in the plots. 
\index{resistant}
There is some suggestion of non-constant variance in the plots but this will be explored
further below when we can remove the difference in the means and combine all
the residuals together.
\index{residuals}
There might be some skew in the responses in some of
the groups but there are only 10 observations per group so 
visual evidence of skew in the boxplots and beanplots 
could be generated by impacts of very few of the observations.

\indent Now we can apply our 6+ steps for performing a hypothesis test
\index{hypothesis testing}
with these observations. The initial step is deciding on the claim to be assessed and the 
test statistic to use. This is a six group situation with a quantitative 
response, identifying it as a One-Way ANOVA where we want to test a null 
hypothesis that all the groups have the same population mean, at least to 
start. We will use a 5% significance level.



1. **Hypotheses**: $\boldsymbol{H_0: \mu_{\text{OJ}0.5} = \mu_{\text{VC}0.5} = \mu_{\text{OJ}1} = \mu_{\text{VC}1} = \mu_{\text{OJ}2} = \mu_{\text{VC}2}} \textbf{ vs }$ $\boldsymbol{H_A:}\textbf{ Not all } \boldsymbol{\mu_j} \textbf{ equal}$

    * The null hypothesis could also be written in reference-coding as below
    since OJ.0.5 is chosen as the baseline group (discussed below).

        * $\boldsymbol{H_0:\tau_{\text{VC}0.5}=\tau_{\text{OJ}1}=\tau_{\text{VC}1}=\tau_{\text{OJ}2}=\tau_{\text{VC}2}=0}$

    * The alternative hypothesis can be left a bit less specific: 
    
        * $\boldsymbol{H_A:} \textbf{ Not all } \boldsymbol{\tau_j} \textbf{ equal 0}$
        
2. **Validity conditions**:

    * Independence:
    
        * This is where the separate cages note above is important. Suppose that 
        there were cages that contained multiple animals and they competed for 
        food or could share illness or levels of activity. The animals in one 
        cage might be systematically different from the others and this 
        "clustering" of observations would present a potential violation of the
        independence assumption. 
        
            If the experiment had the animals in separate 
        cages, there is no clear dependency in the design of the study and we can
        assume^[In working with researchers on hundreds of projects, our experience has been that we often require many conversations to discover all the potential sources of issues in data sets, especially related to assessing independence of the observations.] that there is no problem with this assumption. 
        
    * Constant variance:
    
        * As noted above, there is some indication of a difference in the 
        variability among the groups in the boxplots and beanplots but the sample 
        size was small in each group. We need to fit the linear model to get 
        the other diagnostic plots to make an overall assessment. 

        ```{r Figure3-15,fig.cap="Diagnostic plots for the odontoblast growth model.",fig.height=5,fig.width=10}
        m2 <- lm(len~Treat, data=ToothGrowth)
        par(mfrow=c(2,2))
        plot(m2,pch=16)
        ```
            
        * The Residuals vs Fitted panel in Figure \@ref(fig:Figure3-15) shows some 
        difference in the spreads but the spread is not that different between the groups. 
        
        * The Scale-Location plot also shows just a little less variability in the group 
        with the smallest fitted value but the spread of the groups looks fairly similar in
        this alternative scaling.
        
        * Put together, the evidence for non-constant variance is not that strong 
        and we can proceed comfortably that there is at least not a major problem 
        with this assumption. 
        
    * Normality of residuals: \index{residuals!normality of}
    
        * The Normal Q-Q plot shows a small deviation in the lower tail but nothing that 
        we wouldn't expect from a normal distribution. So there is no evidence of a problem 
        with the normality assumption in the upper right panel of Figure \@ref(fig:Figure3-15). 
        
3. **Calculate the test statistic**:

    * The ANOVA table for our model follows, providing an $F$-statistic of 41.557:
    
    ```{r}
    anova(m2)
    ```


4. **Find the p-value**:\index{p-value}

    * There are two options here, especially since it seems that our assumptions about 
    variance and normality are not violated (note that we do not say "met" -- we just have 
    no clear evidence against them). The parametric and nonparametric approaches should 
    provide similar results here. 
    
    * The parametric approach is easiest -- the p-value comes from the previous 
    ANOVA table as ``< 2e-16``. First, note that this is in scientific notation 
    that is a compact way of saying that the p-value here is $2.2*10^{-16}$ or
    0.00000000000000022.\index{p-value!zero}
    When you see ``2.2e-16`` in R output, it also means
    that the calculation is at the numerical precision limits of the computer. 
    What R is really trying to report is that this is a very small number. 
    **When you encounter p-values that are smaller than 0.0001, you should just
    report that the p-value < 0.0001.** Do not report that it is 0 as this gives 
    the false impression that there is no chance of the result occurring when 
    it is just a really small probability. This p-value came from an $F(5,54)$
    distribution (this is the distribution of the test statistic if the null hypothesis
    is true). 
    
    * The nonparametric approach is not too hard so we can compare the two approaches here  as well:
    
    
    ```{r Figure3-16, fig.cap="Histogram and density curve of permutation distribution for $F$-statistic for odontoblast growth data. Observed test statistic in bold, vertical line at 41.56."}
    Tobs <- anova(lm(len~Treat, data=ToothGrowth))[1,4]; Tobs
    par(mfrow=c(1,2))
    B <- 1000
    Tstar <- matrix(NA, nrow=B)
    for (b in (1:B)){
      Tstar[b] <- anova(lm(len~shuffle(Treat), data=ToothGrowth))[1,4]
    }
    pdata(Tstar, Tobs, lower.tail=F)
    hist(Tstar, xlim=c(0,Tobs+3))
    abline(v=Tobs, col="red", lwd=3)
    plot(density(Tstar), xlim=c(0,Tobs+3), main="Density curve of Tstar")
    abline(v=Tobs, col="red", lwd=3)
    ```
    
    * **The permutation p-value was reported as 0.\index{p-value!zero}
    This should be reported as 
    p-value < 0.001** since we did 1,000 permutations and found that none of the
    permuted $F$-statistics, $F^*$, were larger than the observed $F$-statistic
    of 41.56. The permuted results do not exceed 6 as seen in Figure 
    \@ref(fig:Figure3-16), so the observed result is *really unusual* relative 
    to the null hypothesis. As suggested previously, the parametric and
    nonparametric approaches should be similar here and they were. 
    
5. **Make a decision**:

    * Reject $H_0$ since the p-value is very small. 
    
6. **Write a conclusion and do scope of inference**:

    * There is strong evidence that the different treatments 
    (combinations of OJ/VC and dosage levels) **cause some** difference in the  **true**
    mean odontoblast growth for **these** guinea pigs. 
    
        * We can make the causal statement of the treatment causing differences 
        because the treatments were randomly assigned but these inferences only 
        apply to these guinea pigs since they were not randomly selected from a 
        larger population. 
        
        * Remember that we are making inferences to the population or true 
        means and not the sample means and want to make that clear in any 
        conclusion. When there is not a random sample from a population it is
        more natural to discuss the true means since we can't extend to the 
        population values. 
        
        * The alternative is that there is some difference in the true means 
        -- be sure to make the wording clear that you aren't saying that all 
        the means differ. In fact, if you look back at Figure 
        \@ref(fig:Figure3-14), the means for the 2 mg dosages look almost the 
        same so we will have a tough time arguing that all groups differ. The 
        $F$-test is about finding evidence of some difference *somewhere* among the true means. The next section will
        provide some additional tools to get more specific about the source of 
        those detected differences and allow us to get at estimates of the differences we observed to complete our interpretation. 
        
\indent Before we leave this example, we should revisit our model estimates and 
interpretations. The default model parameterization uses reference-coding. 
Running the model ``summary`` function on ``m2`` provides the estimated 
coefficients:

```{r}
summary(m2)$coefficients
```
    
For some practice with the reference coding used in these models, let's find 
the estimates for observations for a couple of the groups. To work with the 
parameters, you need to start with determining the baseline category that was 
used by considering which level is not displayed in the output. The 
``levels`` function can list the groups in a categorical variable and their 
coding in the data set. The first level is usually the baseline category but 
you should check this in the model summary as well.

```{r}
levels(ToothGrowth$Treat)
```

There is a ``VC.0.5`` in the second row of the model summary, but there is no row for 
``0J.0.5`` and so this must be the baseline category. That means that the fitted value 
or model estimate for the OJ at 0.5 mg/day group is the same as the ``(Intercept)`` row 
or $\hat{\alpha}$, estimating a mean tooth growth of 13.23 microns when the pigs get OJ 
at a 0.5 mg/day dosage level. You should always start with working on the baseline level 
in a reference-coded model. To get estimates for any other group, then you can use the 
``(Intercept)`` estimate and add the deviation for the group of interest. For 
``VC.0.5``, the estimated mean tooth growth is 
$\hat{\alpha} + \hat{\tau}_2 = \hat{\alpha} + \hat{\tau}_{\text{VC}0.5}=13.23 + (-5.25)=7.98$
microns. It is also potentially interesting to directly interpret the estimated difference 
(or deviation) between ``OJ.0.5`` (the baseline) and ``VC.0.5`` (group 2) that 
is $\hat{\tau}_{\text{VC}0.5}= -5.25$: we estimate that the mean tooth growth in
``VC.0.5`` is 5.25 microns shorter than it is in ``OJ.0.5``. This and many other 
direct comparisons of groups are likely of interest to researchers involved in 
studying the impacts of these supplements on tooth growth and the next section 
will show us how to do that (correctly!).

\indent The reference-coding is still going to feel a little uncomfortable so the comparison
to the cell-means model and exploring the effect plot can help to reinforce
that both models patch together the same estimated means for each group. For
example, we can find our estimate of 7.98 microns for the VC0.5 group in the
output and Figure \@ref(fig:Figure3-17). Also note that Figure 
\@ref(fig:Figure3-17) is the same whether you plot the results from 
``m2`` or ``m3``.

```{r}
m3 <- lm(len~Treat-1, data=ToothGrowth)
summary(m3)
```

(ref:fig3-17) Effect plot of the One-Way ANOVA model for the odontoblast growth data.

```{r Figure3-17, fig.cap="(ref:fig3-17)",fig.height=3,fig.width=4,echo=F}
par(mfrow=c(1,1))
plot(allEffects(m2))
```

```{r eval=F}
plot(allEffects(m2))
```

\sectionmark{Multiple (pair-wise) comparisons using Tukey's HSD and CLD}

## Multiple (pair-wise) comparisons using Tukey's HSD and the compact letter display {#section3-6}

\sectionmark{Multiple (pair-wise) comparisons using Tukey's HSD and CLD}

With evidence that the true means are likely not all equal, many researchers 
want to know which groups show evidence of differing from one another. This 
provides information on the source of the overall difference that was 
detected and detailed information on which groups differed from one another. 
Because this is a shot-gun/unfocused sort of approach, some people think it 
is an over-used procedure. Others feel that
it is an important method of addressing detailed questions about group
comparisons in a valid and safe way. For example, we might want to know if OJ is
different from VC *at the 0.5 mg/day* dosage level and these methods will allow 
us to get an answer to this sort of question.
It also will test for differences between the OJ.0.5 and VC.2 groups
and every other pair of levels that you can construct. This method actually
takes us back to the methods in Chapter \@ref(chapter2) where we compared the means of two
groups except that we need to deal with potentially many pair-wise comparisons, 
making an adjustment to account for that inflation in Type I errors
\index{type I error}
that occurs due to many tests being performed at the same time. 
There are many different statistical
methods to make all the pair-wise comparisons, but we will employ the most
commonly used one, called ***Tukey's Honest Significant Difference*** (Tukey's HSD) 
method^[When this procedure is used with unequal group sizes it is also sometimes
called Tukey-Kramer's method.]. The name suggests that not using it could lead to a 
dishonest answer and that it will give you an honest result. It is more that if you don't 
do some sort of correction for all the tests you are performing, you might find some ***spurious***^[We often use "spurious" to describe falsely rejected null hypotheses, but 
they are also called false detections.] results. There are other methods that could be used 
to do a similar correction and also provide "honest" inferences; we are just going to learn 
one of them. 

\indent Generally, the challenge in this situation is that if you perform many tests at the same time (instead of just one test), you inflate the 
Type I error rate.
\index{type I error}
We can define the ***family-wise error rate***
as the probability that at least one error is made on a set of tests or, more
compactly, Pr(At least 1 error is made) where Pr() is the probability of an
event occurring. The family-wise error is meant to capture the overall
situation in terms of measuring the likelihood of making a mistake if we
consider many tests, each with some chance of making their own mistake, and
focus on how often we make at least one error when we do many tests. A quick
probability calculation shows the magnitude of the problem. If we start with a
5% significance level test, then Pr(Type I error on one test) =0.05 and the Pr(no
errors made on one test) =0.95, by definition. This is our standard hypothesis
testing situation. Now, suppose we have $m$ independent tests, then 

$$\begin{array}{ll}
& \text{Pr(make at least 1 Type I error given all null hypotheses are true)} \\
& = 1 - \text{Pr(no errors made)} \\
& = 1 - 0.95^m.
\end{array}$$

Figure \@ref(fig:Figure3-18) shows how the probability of having at least one 
false detection grows rapidly with the number of tests, $m$. The plot stops at 100
tests since it is effectively a 100% chance of at least one false detection. 
It might seem like doing 100 tests is a lot, but in Genetics research it is 
possible to consider situations where millions of tests are
considered so these are real issues to be concerned about in many situations.
Researchers want to make sure that when they report a "significant" result that
it is really likely to be a real result and will show up as a difference in the 
next data set they collect^[Some researchers are now collecting multiple data 
sets to use in a single study and using one data set to identify interesting 
results and then using a validation or test data set that they withheld from 
initial analysis to try to verify that the first results are also present in that 
second data set. This also has problems but the only way to develop an understanding of a process is to look across a suite of studies and learn from that accumulation of evidence.]. 

(ref:fig3-18) Plot of family-wise error rate as the number of tests performed 
increases. Dashed line indicates 0.05. 

```{r Figure3-18, fig.cap="(ref:fig3-18)",echo=F,warning=F,message=F}
m <- 1:100
pr <- 1-.95^m
plot(m,pr,type="l",col="blue",xlab="# of Tests",ylab="Probability of at least one Type I error",lwd=2)
abline(h=.05,lty=2,col="red")
```

\indent In pair-wise comparisons between all the pairs of means in a One-Way ANOVA, the number of 
tests is based on the number of pairs. We can calculate the number of tests using 
$J$ choose 2, $\begin{pmatrix}J\\2\end{pmatrix}$, to get the number of unique pairs of 
size 2 that we can make out of $J$ individual treatment levels. We don't need to 
explore the combinatorics formula for this, as the ``choose`` function in R can give us the 
answers:

\newpage

```{r}
choose(3,2)
choose(4,2)
choose(5,2)
choose(6,2)
```

So if you have three groups (prisoner rating study), there are 3 unique pairs to 
compare. For six groups, like in the guinea pig study, we have to consider 15 tests  to compare all the unique pairs of groups. 15 tests seems like enough that we 
should be worried about inflated family-wise error rates. Fortunately, the 
Tukey's HSD method controls the family-wise error rate at your specified level 
(say 0.05) across any number of pair-wise comparisons. This means that the 
overall rate of at least one Type I error is controlled at the specified 
significance level, often 5%. To do this, each test must use a slightly more 
conservative cut-off than if just one test is performed and the procedure helps 
us figure out how much more conservative we need to be. 

\indent Tukey's HSD starts with focusing on the difference between the groups with the 
largest and smallest means ($\bar{y}_{max}-\bar{y}_{min}$). If 
$(\bar{y}_{max}-\bar{y}_{min}) \le \text{Margin of Error}$
for the difference in the means, then all other pairwise differences, say
$\vert \bar{y}_j - \bar{y}_{j'}\vert$, for two groups $j$ and $j'$, will be less
than or equal to that margin of error. This also means that any confidence
intervals for any difference in the means will contain 0. Tukey's HSD selects a
critical value so that ($\bar{y}_{max}-\bar{y}_{min}$) will be less than the margin of 
error in 95% of data sets drawn from populations with a common mean. This implies
that in 95% of data sets in which all the population means are the same, all
confidence intervals for differences in pairs of means will contain 0. Tukey's
HSD provides confidence intervals for the difference in true means between
groups $j$ and $j'$, $\mu_j-\mu_{j'}$, for all pairs where $j \ne j'$, using

$$(\bar{y}_j - \bar{y}_{j'}) \mp \frac{q^*}{\sqrt{2}}\sqrt{\text{MS}_E\left(\frac{1}{n_j}+
\frac{1}{n_{j'}}\right)}$$

where 
$\frac{q^*}{\sqrt{2}}\sqrt{\text{MS}_E\left(\frac{1}{n_j}+\frac{1}{n_{j'}}\right)}$ 
is the margin of error for the intervals. The distribution used
to find the multiplier, $q^*$, for the confidence intervals is available in the 
``qtukey`` function and generally provides a slightly larger multiplier than the 
regular $t^*$ from our two-sample $t$-based confidence interval discussed in
Chapter \@ref(chapter2). The formula otherwise is very similar to the one used in Chapter \@ref(chapter2) with the SE for the difference in the means based on a measure of residual variance (here $MS_E$) times $\left(\frac{1}{n_j}+\frac{1}{n_{j'}}\right)$ which weights the results based on the relative sample sizes in the groups. 

\indent We will use the ``confint``, ``cld``, and ``plot`` functions 
applied to output from the ``glht`` function (all from the ``multcomp`` package; 
@Hothorn2008, [@R-multcomp]) to easily get the required comparisons from our 
ANOVA model. Unfortunately, its code format is a little complicated -- but there are 
just two places to modify the code, by including the model name and after ``mcp`` 
(stands for *multiple comparisons*) in the ``linfct`` option, you need to include the
explanatory variable name as ``VARIABLENAME="Tukey"``. The last part is to get the 
Tukey HSD multiple comparisons run on our explanatory variable. Once we obtain the
intervals, we can use them to test $H_0: \mu_j = \mu_{j'} \text{ vs } H_A: \mu_j \ne \mu_{j'}$ 
by assessing whether 0 is in the confidence interval for each pair. If 0 is in the 
interval, then there is no evidence of a difference for that pair. If 0 is not in the 
interval, then we reject $H_0$ and have evidence *at the specified family-wise significance
level* of a difference for that pair. You will see a switch to using the
word "detection" to describe rejected null hypotheses of no difference as it
can help to compactly write up these results. The following code provides the numerical
and graphical^[The plot of results usually contains all the labels of groups but if the labels are long or there many groups, sometimes the row labels are hard to see  even with re-sizing the plot to make it taller in RStudio. The numerical output is
useful as a guide to help you read the plot.] results of applying Tukey's HSD
to the linear model for the Guinea Pig data:

(ref:fig3-19) Graphical display of pair-wise comparisons from Tukey's HSD for the 
Guinea Pig data. Any confidence intervals that do not contain 0 provide evidence 
of a difference in the pair of groups.

```{r Figure3-19, fig.cap="(ref:fig3-19)", fig.height=5,fig.width=10}
require(multcomp)
Tm2 <- glht(m2, linfct = mcp(Treat = "Tukey"))
confint(Tm2)
old.par <- par(mai=c(1,2,1,1)) #Makes room on the plot for the group names
plot(Tm2)
```

\indent Figure \@ref(fig:Figure3-19) contains confidence intervals for the difference in 
the means for all 15 pairs of groups. For example, the first row in the plot contains 
the confidence interval for comparing VC.0.5 and OJ.0.5 (VC.0.5 **minus** OJ.0.5). In the numerical output, you can find that this 95% 
family-wise confidence interval goes from -10.05 to -0.45 microns (``lwr`` and 
``upr`` in the numerical output provide the CI endpoints). This interval does not 
contain 0 since its upper end point is -0.45 microns and so we can now say that 
there is evidence that OJ and VC have different true mean growth rates at the 0.5 mg 
dosage level. We can go further and say that we are 95% confident that the difference 
in the true mean tooth growth between VC.0.5 and OJ.0.5 (VC.0.5-OJ.0.5) is between 
-10.05 and -0.45 microns, after adjusting for comparing all the pairs of groups. 
But there are fourteen more similar intervals...

\indent If you put all these pair-wise tests together, you can generate an overall 
interpretation of Tukey's HSD results that discusses sets of groups that 
are not detectably different from one another and those groups that were 
distinguished from other sets of groups. To do this, start with listing 
out the groups that are not detectably different (CIs contain 0), which, 
here, only occurs for four of the pairs. The CIs that contain 0 are for the 
pairs VC.1 and OJ.0.5, OJ.2 and OJ.1, VC.2 and OJ.1, and, finally, VC.2 and 
OJ.2. So VC.2, OJ.1, and OJ.2 are all not detectably different from each 
other and VC.1 and OJ.0.5 are also not detectably different. If you look carefully,  VC.0.5 is detected as different from every other group. So there are basically 
three sets of groups that can be grouped together as "similar": VC.2, OJ.1, 
and OJ.2; VC.1 and OJ.0.5; and VC.0.5. Sometimes groups overlap with some 
levels not being detectably different from other levels that belong to 
different groups and the story is not as clear as it is in this case. An 
example of this sort of overlap is seen in the next section. 

\indent There is a method that many researchers use to more efficiently generate and 
report these sorts of results that is called a ***compact letter display*** 
(CLD, @Piepho2004)^[Note that this method is implemented slightly differently than we explain here in some software packages so if you see this in a journal article, read the discussion carefully.]. The ``cld`` function can be applied to the results from 
``glht`` to generate the CLD that we can use to provide a "simple" summary of 
the sets of groups. In this discussion, we define a **set as a union of different
groups that can contain one or more members** and the member of these groups are 
the different treatment levels. 

```{r}
cld(Tm2)
```

Groups with the same letter are not detectably different (are in the same set) 
and groups that are detectably different get different letters (are in 
different sets). Groups can have more than one letter to reflect "overlap" 
between the sets of groups and sometimes a set of groups contains only a 
single treatment level (VC.0.5 is a set of size 1). Note
that if the groups have the same letter, this does not mean they are the same, 
just that there is **insufficient evidence to declare a difference for that pair**. If we consider 
the previous output for the CLD, the "a" set contains VC.0.5, the "b" set contains 
OJ.0.5 and VC.1, and the "c" set contains OJ.1, OJ.2, and VC.2. These are exactly 
the groups of treatment levels that we obtained by going through all fifteen 
pairwise results. 

\indent One benefit of this work is that the CLD letters can be added to a plot (such as the beanplot) to 
help fully report the results and understand the sorts of differences Tukey's 
HSD detected. The lines with ``text`` in them are involved in placing text on the figure but are 
something you could do in image editing software just as easily. 
Figure \@ref(fig:Figure3-20) enhances the discussion by showing that the 
"`r colFmtBold("a","blue")`" group with VC.0.5 had the lowest average tooth 
growth, the "`r colFmtBold("b","red")`" group had intermediate tooth growth
for treatments OJ.0.5 and VC.1, and the highest growth rates came from
OJ.1, OJ.2, and VC.2. Even though VC.2 had the highest average growth rate, 
we are not able to prove that its true mean is any higher
than the other groups labeled with "`r colFmtBold("c","green")`". Hopefully the 
ease of getting to the story of the Tukey's HSD results from a plot like this 
explains why it is common to report results using these methods instead of 
reporting 15 confidence intervals for all the pair-wise differences. 

(ref:fig3-20) Beanplot of odontoblast growth by group with Tukey's HSD compact 
letter display.

```{r Figure3-20, fig.cap="(ref:fig3-20)",echo=F,fig.width=10,fig.height=6}
old.par <- par(mai=c(1,1,1,1))
beanplot(len~Treat, data=ToothGrowth, log="", col="white", method="jitter")
text(c(2),c(10),"a",col="blue",cex=1.5)
text(c(1,4),c(15,18),"b",col="red",cex=1.5)
text(c(3,5,6),c(25,28,28),"c",col="green",cex=1.5)
```


\indent There are just a couple of other details to mention on this set of methods. First, 
note that we interpret the set of confidence intervals simultaneously: We are
95% confident that **ALL** the intervals contain the respective differences in 
the true means (this is a ***family-wise interpretation***). These intervals are 
adjusted from our regular 2 sample $t$ intervals from Chapter \@ref(chapter2)
to allow this stronger interpretation. Specifically, 
they are wider. Second, if sample sizes are unequal in the groups, Tukey's HSD
is conservative and provides a family-wise error rate that is lower than the
*nominal* (or specified) level. In other words, it fails less often than expected
and the intervals provided are a little wider than needed, containing all the
pairwise differences at higher than the nominal confidence level of (typically)
95%. Third, this is a parametric approach and violations of normality and
constant variance will push the method in the other direction, potentially
making the technique dangerously liberal. Nonparametric approaches to this
problem are also possible, but will not be considered here. 

## Pair-wise comparisons for Prisoner Rating data {#section3-7}

In our previous work with the prisoner rating data, the overall ANOVA test
provided only marginal evidence of some difference in the true means across the
three groups with a p-value=0.067. Tukey's HSD does not require you 
to find a small p-value from your overall $F$-test to employ the methods 
but if you apply it to situations with p-values larger than your 
*a priori* significance level,
you are unlikely to find any pairs that are detected as being different. Some
statisticians suggest that you shouldn't employ follow-up tests such as Tukey's
HSD when there is not sufficient evidence to reject the overall null hypothesis
and would be able to reasonably criticize the following results. But for the
sake of completeness, we can find the pair-wise comparison results at our
typical 95% family-wise confidence level in this situation, with the three
confidence intervals displayed in Figure \@ref(fig:Figure3-21). 

(ref:fig3-21) Tukey's HSD confidence interval results at the 95% family-wise confidence level for the Mock Jury linear model.

```{r Figure3-21,fig.cap="(ref:fig3-21)",fig.height=4,echo=F}
lm2 <- lm(Years~Attr, data=MockJury)
require(multcomp)
Tm2 <- glht(lm2, linfct = mcp(Attr = "Tukey"))
old.par <- par(mai=c(1,2.5,1,1)) #Makes room on the plot for the group names
plot(Tm2)
```

```{r eval=F}
lm2 <- lm(Years~Attr, data=MockJury)
require(multcomp)
Tm2 <- glht(lm2, linfct = mcp(Attr = "Tukey"))
```

\vspace{0.25in}

```{r}
confint(Tm2)
cld(Tm2)
```

```{r eval=F}
old.par <- par(mai=c(1,2.5,1,1)) #Makes room on the plot for the group names
plot(Tm2)
```

(ref:fig3-22) Tukey's HSD 90% family-wise confidence intervals for the Mock Jury linear model.

```{r Figure3-22, fig.cap="(ref:fig3-22)",fig.height=4}
confint(Tm2, level=0.9)
cld(Tm2, level=0.1)
old.par <- par(mai=c(1,2.5,1,1)) #Makes room on the plot for the group names
plot(confint(Tm2, level=.9))
```

\indent At the family-wise 5% significance level, there are no pairs that are detectably different 
-- they all get the same letter of "a". Now we will produce results for the reader that
thought a 10% significance was suitable for this application before seeing any
of the results. We just need to change the confidence level or significance
level that the CIs or tests are produced with inside the functions. For the ``confint``
function, the ``level`` option is the confidence level and for the ``cld``, it is the 
family-wise significance level. Note that 90% confidence corresponds to a 10% 
significance level. 

```{r}
confint(Tm2, level=0.9)
```


```{r}
cld(Tm2, level=0.1)
```

```{r eval = F}
old.par <- par(mai=c(1,2.5,1,1)) #Makes room on the plot for the group names
plot(confint(Tm2, level=.9))
```

(ref:fig3-23) Beanplot of sentences with compact letter display results from 10%
family-wise significance level Tukey's HSD. *Average* and *Unattractive* picture
groups are detected as being different and are displayed as belonging to
different groups. *Beautiful* picture responses are not detected as different
from the other two groups.

```{r Figure3-23, fig.cap="(ref:fig3-23)"}
old.par <- par(mai=c(0.5,1,1,1))
beanplot(Years~Attr, data=MockJury, log="", col="white", method="jitter")
text(c(1), c(5.3),"ab", col="blue", cex=1.5)
text(c(2), c(5.1),"a", col="green", cex=1.5)
text(c(3), c(6.8),"b", col="red", cex=1.5)
```

With family-wise 10% significance and 90% confidence levels, the 
*Unattractive* and *Average* picture groups are detected as being different but
the *Average* group is not detected as different from *Beautiful* and 
*Beautiful* is not detected to be different from *Unattractive*. This leaves the
"overlap" of groups across the sets of groups that was noted earlier. The 
*Beautiful* level is not detected as being dissimilar from levels in two 
different sets and so gets two different letters. 

\indent The beanplot (Figure \@ref(fig:Figure3-23)) helps to clarify some of the reasons
for this set of results. The detection of a difference between *Average* and
*Unattractive* just barely occurs and the mean for *Beautiful* is between the 
other two so it ends up not being detectably different from either one. This 
sort of overlap is actually a fairly common occurrence in these sorts of 
situations so be prepared a mixed set of letters for some levels. 

```{r eval = F}
old.par <- par(mai=c(0.5,1,1,1))
beanplot(Years~Attr, data=MockJury, log="", col="white", method="jitter")
text(c(1), c(5.3),"ab", col="blue", cex=1.5)
text(c(2), c(5.1),"a", col="green", cex=1.5)
text(c(3), c(6.8),"b", col="red", cex=1.5)
```

## Chapter summary {#section3-8}

In this chapter, we explored methods for comparing a quantitative response across 
$J$ groups ($J \ge 2$), with what is called the One-Way ANOVA procedure. The initial 
test is based on assessing evidence against a null hypothesis of no difference in the true means for the $J$ groups. There are two different methods for estimating these One-Way ANOVA models: 
the cell-means model and the reference-coded versions of the model. 
There are times when either model will be preferred, but for the rest of the text, 
the reference coding is used (sorry!). The ANOVA $F$-statistic, 
often presented with 
underlying information in the ANOVA table,
\index{ANOVA table}
provides a method of assessing evidence 
against the null hypothesis either using permutations or via the $F$-distribution. 
Pair-wise comparisons using Tukey's HSD provide a method for comparing all the groups 
and are a nice complement to the overall ANOVA results. A compact letter display was 
shown that enhanced the interpretation of Tukey's HSD result. 

\indent In the guinea pig example, we are left with some lingering questions based on these 
results. It appears that the effect of *dosage* changes as a function of the 
*delivery method* (OJ, VC) because the size of the differences between OJ and VC
change for different dosages. These methods can't directly assess the question
of whether the effect of delivery method is the same or not across the
different dosages. In Chapter \@ref(chapter4), the two variables, *Dosage* and 
*Delivery method* are modeled as two separate variables so we can consider their 
effects both separately and together. This allows more refined hypotheses, such as 
*Is the effect of delivery method the same for all dosages?*, to be tested. This 
will introduce new models and methods for analyzing data where there are two 
factors as explanatory variables in a model for a quantitative response variable 
in what is called the Two-Way ANOVA. 

## Summary of important R code {#section3-9}

The main components of R code used in this chapter follow with components to 
modify in lighter and/or ALL CAPS text, remembering that any R packages mentioned 
need to be installed and loaded for this code to have a chance of working:

* **`r colFmt("MODELNAME","red")` <- lm(`r colFmt("Y","red")`~`r colFmt("X","red")`,
data=`r colFmt("DATASETNAME","red")`)**

    * Probably the most frequently used command in R. 
    
    * Here it is used to fit the reference-coded One-Way ANOVA model with Y as the 
    response variable and X as the grouping variable, storing the estimated model 
    object in MODELNAME. \index{\texttt{lm()}|textbf}

* **`r colFmt("MODELNAME","red")` <- lm(`r colFmt("Y","red")`~`r colFmt("X","red")`-1,
data=`r colFmt("DATASETNAME","red")`)**

    * Fits the cell means version of the One-Way ANOVA model. 

* **summary(`r colFmt("MODELNAME","red")`)**

    * Generates model summary information including the estimated model coefficients, 
    SEs, t-tests, and p-values. 

* **anova(`r colFmt("MODELNAME","red")`)**

    * Generates the ANOVA table but **must only be run on the 
    reference-coded version of the model**. \index{\texttt{anova()}|textbf}
    
    * Results are incorrect if run on the cell-means model since the reduced model 
    under the null is that the mean of all the observations is 0!

* **pf(`r colFmt("FSTATISTIC","red")`, df1=`r colFmt("NUMDF","red")`,
df2=`r colFmt("DENOMDF","red")`, lower.tail=F)**

    * Finds the p-value for an observed $F$-statistic with NUMDF and DENOMDF degrees 
    of freedom. \index{\texttt{pf()}|textbf}

* **par(mfrow=c(2,2)); plot(`r colFmt("MODELNAME","red")`)**

    * Generates four diagnostic plots including the Residuals vs Fitted and 
    Normal Q-Q plot. 
    
* **plot(allEffects(`r colFmt("MODELNAME","red")`))**
    
    * Requires the ``effects`` package be loaded. 

    * Plots the estimated model component. \index{\texttt{allEffects()}|textbf}
    
* **Tm2 <- glht(`r colFmt("MODELNAME","red")`, linfct=mcp(`r colFmt("X","red")`="Tukey"));
confint(Tm2); plot(Tm2); cld(Tm2)**

    * Requires the ``multcomp`` package to be installed and loaded.
    
    * Can only be run on the reference-coded version of the model. 
    
    * Generates the text output and plot for Tukey's HSD as well as the compact 
    letter display. 

## Practice problems {#section3-10}

For these practice problems, you will work with the cholesterol data set from 
the ``multcomp`` package that was used to generate the Tukey's HSD results. To 
load the data set and learn more about the study, use the following code:

```
require(multcomp)
data(cholesterol)
require(tibble)
cholesterol <- as.tibble(cholesterol)
help(cholesterol)
```

3.1. Graphically explore the differences in the changes in Cholesterol levels for
the five levels using boxplots and beanplots. 

3.2. Is the design balanced? \index{balanced}

3.3. Complete all 6+ steps of the hypothesis test using the parametric $F$-test, 
reporting the ANOVA table and the distribution of the test statistic under the null. 

3.4. Discuss the scope of inference using the information that the treatment levels
were randomly assigned to volunteers in the study. 

3.5. Generate the permutation distribution and find the p-value. Compare the parametric
p-value to the permutation test results. 

3.6. Perform Tukey's HSD on the data set. Discuss the results -- which pairs were
detected as different and which were not? Bigger reductions in cholesterol are
good, so are there any levels you would recommend or that might provide similar
reductions?

3.7. Find and interpret the CLD and compare that to your interpretation of results 
from 3.6.

